The PMA Phorbol Ester Tumor Promoter Increases Canonical Wnt Signaling Via Macropinocytosis

Version 1

Abstract

Summary

Activation of the Wnt pathway lies at the core of many human cancers. Interestingly, Wnt, cell adhesion, and macropinocytosis are often active in the same processes, and understanding how Wnt signaling and membrane trafficking cooperate should improve our understanding of embryonic development and cancer. Here we show that a macropinocytosis activator, the tumor promoter Phorbol 12-myristate 13-acetate (PMA), enhances Wnt signaling. Experiments using the Xenopus embryo as an in vivo model showed marked cooperation between the PMA phorbol ester and Wnt signaling, which was blocked by inhibitors of macropinocytosis, Rac1 activity, and lysosome acidification. The crosstalk between canonical Wnt, the Protein Kinase C (PKC) pathway, focal adhesions, lysosomes, and macropinocytosis suggests possible therapeutic targets for cancer progression in Wnt-driven cancers.

Introduction

Cancer progression is a process during which mutations in various signaling pathways accumulate in the cell leading to uncontrolled growth (Kinzler and Vogelstein, 1996; Hanahan and Weinberg, 2011). Cancer progression is also influenced by tumor promoters, substances that do not mutate the DNA by themselves but nevertheless facilitate cancer development. In 1941 I. Berenblum and Peyton Rous found that inflammatory agents such as croton oil or turpentine could induce skin cancer when applied repeatedly to mouse or rabbit skin, but only when an initiator mutagen such as benzopyrene or anthracene had been previously painted on that patch of skin (Berenblum, 1941; Rous and Kidd, 1941). The active agent in croton oil was later purified and found to be the phorbol ester Phorbol-12-myristate-13-acetate (PMA), also known as 12-O-tetradecanoylphorbol-13-acetate (TPA). The molecular mechanism by which PMA promotes cancer is incompletely understood but involves the activation of Protein Kinase C (PKC) in the plasma membrane, where PMA mimics the endogenous second messenger Diacylglycerol (DAG) (Nishizuka, 1984). Tumor promoters cause chronic inflammation and significantly contribute to tumorigenesis (Weinberg, 2007).

The canonical Wnt/β-catenin pathway is a major driver of cancer and plays a crucial role in embryonic development (MacDonald et al., 2009; Nusse and Clevers, 2017). In the case of colon cancer, the great majority of tumors are initiated by mutation of the tumor suppressor Adenomatous Polyposis Coli (APC), which is a component of the destruction complex that normally degrades the transcriptional regulator β-catenin, leading to its accumulation in the nucleus and benign polyp formation (Kinzler and Vogelstein, 1996; Segditsas and Tomlinson, 2006). Activation of the Wnt/β-catenin pathway is also the driving force in many other cancers (Galluzzi et al., 2019). A key regulator of Wnt signaling is Glycogen Synthetase 3 (GSK3), which phosphorylates β-catenin, leading to the formation of a phosphodegron that triggers its degradation in the proteasome. Upon activation of the Wnt receptors, GSK3, together with the destruction complex, is sequestered into multivesicular bodies (MVBs), also known as the late endosomes (Taelman et al., 2010; Vinyoles et al., 2014). The translocation of GSK3 from the cytosol into MVBs/lysosomes leads to the stabilization of many other cellular proteins in addition to β-catenin, in a process designated Wnt-STOP (Wnt-Stabilization of Proteins) (Acebron et al., 2014; Albrecht et al., 2021).

The many GSK3 phosphorylation targets stabilized by Wnt signaling include regulators of macropinocytosis such as Ras, Rac1, and Pak1 (Taelman et al., 2010; Albrecht et al., 2020; Jeong et al., 2012). A recent advance was the realization that Wnt signaling requires macropinocytosis (i.e., cell drinking) in order to traffic activated Wnt receptors and GSK3 to MVBs and lysosomes (Redelman-Sidi et al., 2018; Tejeda-Muñoz et al., 2019). Wnt-induced macropinocytosis also causes the internalization of focal adhesions and integrins from the cell surface into MVBs (Tejeda et al., 2022). The requirement of macropinocytosis for Wnt signaling is revealed, for example, by the decrease in nuclear β-catenin when colorectal cancer (CRC) cells mutant for APC were treated with macropinocytosis inhibitors such as the derivative of the diuretic Amiloride known as 5-(N-Ethyl-N-isopropyl) Amiloride (EIPA) (Tejeda-Muñoz et al., 2019; Tejeda-Muñoz and De Robertis, 2022a, 2022b;). Macropinocytosis is emerging in the field as a promising target for cancer treatment (Lambies and Commisso, 2022). In addition, it has been found that the macropinocytosis induced by Wnt signaling causes significant lysosome acidification and activation (Albrecht et al., 2020). There is also accumulating evidence showing that the size and number of lysosomes increase during carcinogenesis (Cardone et al., 2005; Kroemer and Jäättelä, 2005; Zhitomirsky and Assaraf, 2016). Lysosomes, previously thought to be the garbage disposal of the cell, have emerged as central organelles in cancer (Kallunki et al., 2012; Kirkegaard and Jäättelä, 2009).

From the considerations above, investigating the intersection of Wnt signaling, macropinocytosis, lysosomes, focal adhesions, and membrane trafficking, is of great interest. While in most cells macropinocytosis requires stimulation of receptor tyrosine kinase (RTK) (Haigler et al., 1979; Yoshida et al., 2018), Ras (Bar-Sagi and Feramisco 1986; Commisso, 2013), or Wnt (Redelman-Sidi et al., 2018; Tejeda et al., 2019), some cells, such as amoebae and macrophages, as have constitutive macropinocytosis. Importantly, Swanson discovered that the phorbol ester PMA stimulated macropinocytosis even in macrophages that were thought to have fully activated constitutive macropinocytosis (Swanson, 1989).

In the present study, we asked whether phorbol ester could interact with canonical Wnt signaling. Using the Xenopus embryo, which provides a premier model system to analyze the Wnt pathway, we found that PMA, which lacked any effects on its own, greatly sensitized the embryo to dorsalization by Wnt signaling, leading to the formation of radial head structures lacking trunks. The dorsalizing effect of PMA on embryos was blocked by the macropinocytosis inhibitor EIPA, the Rac1 inhibitor EHT1864 (an upstream regulator of an upstream regulator of p21-activated kinase 1, Pak1, which is required for actin-mediated macropinocytosis), and the lysosomal V-ATPase inhibitor Bafilomycin A (Baf). Lysosomal activity was specifically required at the 32-cell stage for the activity of the Wnt-mimicking GSK3 inhibitor Lithium Chloride (LiCl). Ectopic axes induced by dominant-negative GSK3 (DN-GSK3-β) or β-catenin mRNAs were blocked by EIPA or Dominant-negative Rab7, indicating a widespread requirement of membrane trafficking for Wnt signaling. This involvement of membrane trafficking in Wnt-driven cancer was supported by immunostaining results of arrayed human CRC histological sections, in which we found that the grade of malignancy increased Pak1, CD63 (an MVB marker), and V0a3 (a V-ATPase subunit that marks lysosomes) levels, while decreasing GSK3. In Xenopus embryos, twinning induced by Xwnt8 mRNA microinjection was blocked by inhibition of Rac1. In cultured CRC SW480 cells reconstituted with APC, the focal adhesion marker Tes was cytoplasmic, while the same cells mutant for APC (in which the Wnt pathway is activated) Tes and FAK (Focal Adhesion Kinase) were nuclear. The results suggest that macropinocytosis and membrane trafficking play an important role in embryogenesis and cancer progression.

Results

The Phorbol Ester PMA Potentiates Wnt-like Signals in Xenopus embryos Via Macropinocytosis and Lysosomal Acidification

Wnt is one of the earliest signals in the embryo and it induces the formation of the primary dorsal-ventral and head-to-tail axes, organogenesis, asymmetrical stem cell division, and regeneration (Loh et al., 2016; Niehrs, 2012). It is known that β-catenin is required for Wnt signaling activation on the dorsal side during early cleavage. Recently, we reported that inhibition of macropinocytosis with EIPA, or of lysosomal acidification with Baf, ventralized endogenous early dorsal axis formation, leading in particular to reduced head and brain structures (Tejeda-Munoz and De Robertis, 2022a). Activation of lysosomes on the dorsal side was detectable already at the 64-cell stage (Tejeda-Munoz and De Robertis, 2022a). To detect macropinocytosis directly, we now injected the macropinocytosis marker TMR-dextran 70kDa, which has a hydrated diameter of more than 200 nM and is diagnostic of large endocytic cups (Commisso et al., 2013, 2014).

We found that microinjection of TMR-dextran into the gastrula cavity showed stronger macropinocytosis just above the dorsal lip of the blastopore (Fig. 1A and B, indicated with an asterisk). While macropinocytosis probably also occurs at earlier stages, the blastocele cavity is not large enough during the cleavage stages to perform this microininjection experiment.

The Phorbol Ester PMA Synergizes with GSK3 Inhibition and this Cooperation Requires Macropinocytosis and Lysosome Acidification (A-A’) Albino mbryos injected into the gastrula cavity (40 nl at 1 mM) with the macropinocytosis marker TMR-dextran 70kDa were fixed and bisected at gastrula stage 11. Note that dextran macropinocytosis is stronger in the dorsal part of the embryo (asterisk; dorsal blastopore indicated by arrowhead). (C) PMA increases the effects of LiCl in Wnt signaling in early embryos. qRT-PCR analysis at blastula stage 9.5 of the Wnt target genes Siamois and Xnr3 normalized for Ornithine decarboxylase (ODC) transcripts. Embryos received a single injection of 4 nl of LiCl (300 mM with or without PMA (500 nM). (D) Uninjected control embryo at stage 24. (E) LiCl (4nl 300 mM, 1 x ventral) dorsalized the embryo only slightly. (F) EIPA (1 mM, 1x ventral) alone did not produce a distinctive phenotype at this concentration. (G) The vacuolar ATPase (V-ATPase) inhibitor Bafilomycin A1 (Baf) (incubation with 5 μM for 7 minutes at 32 cells) ventralized embryos. (H) A single injection of PMA (500 nM) did not show any phenotypic effect. (I) PMA and LiCl strongly synergized, resulting in embryos with radial heads. (J) Coinjection of the EIPA macropinocytosis inhibitor blocked dorsalization caused by LiCl plus PMA. (K) Incubation with Baf lysosomal acidification inhibitor blocked dorsalization caused by LiCl plus PMA. Number of embryos or hemisections analyzed were as follows: A = 150, 100%; A’ = 160, 97% with fluorescence; five independent experiments; D = 70, 100%; E = 80; 98%; F = 75, 94%; G = 82, 92%; H = 76, 100%; I = 112, 98% J = 75, 93%; K = 85, 98%. Scale bars, 500 μm; Error bars denote SEM (n ≥ 3) (**P < 0.01). Also see Figure S1.

To test whether increasing macropinocytosis affected embryonic dorsal patterning, we used microinjections of PMA, which is known to increase macropinocytosis in macrophages (Swanson, 1989). We used a sensitized system in which a small amount of the GSK3 inhibitor Lithium chloride (LiCl) (4 nl at 300 mM in a single ventral injection at 4-cells) causes only a small dorsalization with slightly enlarged heads (Kao et al., 1986; Tejeda-Munoz and De Robertis, 2022a). The early Wnt signal can be detected at blastula (stage 9.5, just after the start of zygotic transcription) by measuring transcripts of the Wnt target genes Siamois and Xnr3 by qRT-PCR (Tejeda-Munoz and De Robertis, 2022a). As shown in Figure 1C, PMA alone (4 nl at 500 nM) was without any effect, LiCl increased the Wnt target genes, and the combination of PMA and LiCl synergized significantly, elevating Siamois and Xnr3 transcript levels.

The dorsalizing effects of PMA were most striking when phenotypes were analyzed at the early tailbud stage (Figure 1D-K). LiCl alone expanded head structures only slightly and PMA alone was without effect (Figures 1D, E, and H). However, the combination of PMA and LiCl lead to radially dorsalized embryos consisting of head structures lacking trunks (compare Figure 1E to 1I). This striking phenotype was eliminated by co-injection of EIPA or Baf, indicating that macropinocytosis and lysosomal acidification are required for the effects of PMA (Figure 1J and 1K). Injection of Baf alone decreased head and neural development, supporting a requirement for lysosomes for endogenous dorsal axis formation (Figure 1G). In target gene transcription assays, EIPA and Baf also blocked the increase in Siamois and Xnr3 caused by PMA/LiCl (Figure 1C).

Using the BAR (β-catenin activity reporter) assay system (Biechele and Moon, 2008) in a cultured HEK293 permanent BAR/Renilla cell line, PMA synergized with LiCl when assayed both by β-catenin protein levels and BAR luciferase activity (Supplementary Figure S1). PMA alone caused a slight increase in β-catenin protein that lacked statistical significance (Figure S2C and S2I). This is presumably due to a small amount of endogenous Wnt secreted by this cell line (Colozza et al., 2020). However, PMA on its own had no detectable effect on BAR reporter activity (Figure S1J). Importantly, the effect of PMA on the activation of the Wnt pathway by GSK3 inhibition was abrogated by the inhibition of V-ATPase with Baf (Figure S1).

These results in embryos and cultured cells indicate that the tumor promoter PMA potentiates the Wnt signaling pathway and that this cooperation requires macropinocytosis and lysosomal acidification.

Lysosomal Activity is Required at the 32-cell Stage for Early Wnt Activation

In previous work, we had shown that immersion of embryos in the V-ATPase inhibitor Baf at 5 µM for only 7 minutes at the 32-cell stage resulted in macrocephalic embryos lacking cement glands (Tejeda-Muñoz and De Robertis, 2022a). To test whether this ventralization is caused by inhibition of the endogenous Wnt pathway signal, we now followed an initial Baf incubation with a second brief incubation in LiCl while still at the 32-cell stage. LiCl treatment has been shown to induce macropinocytosis in Xenopus embryos (Albrecht et al., 2020). It was found that ventralization by inhibiting V-ATPase was entirely rescued by the subsequent 7 minute pulse of LiCl, resulting in strongly dorsalized tadpoles with large heads (Figure 2A-D). These phenotypes correlate with the activation of Wnt signaling, as indicated by qRT-PCR of Wnt targets at the blastula stage (Figure 2E). In the converse experiment, when embryos were treated with LiCl, washed, and then incubated for 7 minutes in Baf, the effect of lysosomal inhibition was dominant, resulting in ventralized tailbud embryos (Figure 2F-2J). This order-of-addition experiment strongly supports the proposal (Tejeda-Muñoz and De Robertis, 2022a) that lysosomal activity is required for the initial endogenous Wnt signal at the 32-cell stage of development.

Order-of-addition Experiment Showing that Lysosomal Acidification at 32-cell Stage is Required for Early Wnt/β-catenin Activation (A) Untreated embryo. (B) Baf treatment (5 μM for 7 minutes) inhibited endogenous axis formation resulting in ventralized embryos lacking heads at early tailbud. (C) Embryos treated with Baf, washed, and then immersed in 300 mM LiCl an additional 7 minutes formed heads, resulting in dorsalized embryos with enlarged heads and short trunks. Note that all treatments were done at 32-cell stage, which is critical for the early Wnt signal. (D) LiCl treatment alone dorsalized embryos. (E) Quantitative RT-PCR (qPCR) for Wnt target genes Siamois and Xnr3 at blastula, confirming that the phenotypic effects are due to early activation of the Wnt pathway. (F) Untreated embryo. (G) Embryos incubated with LiCl at 32-cell were dorsalized. (H) Embryos first treated with LiCl and then subsequently with Baf resulted in ventralized embryos lacking heads. (I) Baf treatment caused ventralization and small heads. (J) Quantitative RT-PCR for Wnt target genes Siamois and Xnr3 showing that inhibiting V-ATPAse at the 32-cell stage blocks the effect of earlier LiCl treatment. The numbers of embryos analyzed were as follows: A = 115, 100% B = 124, 98% with phenotype; C = 132, 97%; D 99%; F = 132, 100%; G = 135, 99%; H = 126, 97%; I = 129, 98% (Scale bars, 500 μm.). Error bars denote SEM (n ≥ 3) (**P < 0.01).

Wnt Pathway Signaling in Microinjected Embryos or APC Mutant Cells Requires Macropinocytosis and Membrane Trafficking

Next, we tested the requirement for macropinocytosis of different components of the Wnt pathway using the axis duplication assay (McMahon and Moon, 1989). Microinjection of Wnt-mimic dominant-negative GSK3 (DN-GSK3-β) mRNA, a catalytically inactive form of GSK3, induced complete axis duplications, which were blocked by the co-injection of the macropinocytosis inhibitor EIPA (1 mM, 4 nl) (Figure 3A-D). Secondary axis induction by β-catenin mRNA was also abrogated by EIPA, as well as the membrane trafficking inhibitor dominant-negative Rab7 (DN-Rab7) (Figure 3E-H). β-catenin is a transcriptional activator, and one might have not expected a requirement for membrane trafficking. However, axis formation is a multistep process and appears to require macropinocytosis and Rab7 activity in addition to transcriptional activation of Wnt target genes.

Induction of Secondary Axes by the GSK3/β-catenin Pathway Requires Membrane Trafficking (A) Control embryo at early tailbud. (B) Embryo injected with DN-GSK3 (dominant-negative GSK3-β, 150 pg 1 x ventral) mRNA, showing double axes (arrowheads). (C) Injection of EIPA 1 x ventral alone showed no phenotypic effect (4 nl,1 mM). (D) Coinjection of DN-GSK3 and EIPA blocked double axis formation. (E) Uninjected control embryo. (F) Activation of Wnt signaling via injection of β-catenin mRNA (80 pg) induced complete twinned axes (arrowheads). (G) Embryo coinjected with β-catenin mRNA and DN-Rab7 (500 pg) showing that membrane trafficking is required for secondary axis formation. (H) Coinjection of β-catenin mRNA and the macropinocytosis inhibitor EIPA blocks axial duplication. The numbers of embryos analyzed were as follows: A = 62, 100%; B = 75, 94% with double axes; C = 76, 100%; D = 70; 98%; E = 70, 100%; F = 73, 98%; G= 69, 75%; H = 82, 97%; four independent experiments; (Scale bars, 500 μm.). Error bars denote SEM (n ≥ 3) (**P < 0.01).

Macropinocytosis is an actin-driven process that is orchestrated by Pak1 kinase, which in turn is activated by Rac1. Rac1 activity is essential for the formation of lamellipodia and macropinocytic cups (Hall, 1998). EHT1864 (EHT) is a specific inhibitor of Rac1 that has the practical advantage of being water-soluble (Hampsch et al., 2017). When embryos were immersed in EHT at the critical 32-cell stage for 7 minutes (10 mM), head structures were greatly reduced, and ventral posterior structures expanded (Figures 4A and 4B). This was accompanied by a significant increase of transcripts of the ventral markers Sizzled and Vent1 at the blastula stage (Figure 4C).

Rac1 Inhibitor EHT1864 Blocks Wnt Signaling in Xenopus Embryos, and Macropinocytosis Is Required for Nuclear β-catenin Accumulation in CRC Cells (A) Uninjected control embryo. (B) Incubation of the Xenopus embryos with the Rac1 inhibitor EHT at 32-cell stage (7 min, 10 mM) resulted in a ventralized phenotype with a small head in the anterior (A, arrowhead) and expanded ventral structures in the posterior (P). Rac1 activity is required for macropinocytosis, see Figure S2). (C) qRT-PCR of gastrula stage embryos showing increased ventral markers Szl and Vent1 after Rac1 inhibition. (D) Control embryo. (E) Injection of Wnt8 mRNA (2 pg) induces complete duplicated axes (arrows). (F) Injected embryos with EHT (1 mM, 4 nl 1 x ventral) alone showed no phenotypic effect at this concentration. (G) EHT coinjected with Wnt8 mRNA blocked double axis formation. (H) qRT-PCR of Wnt target genes Siamois and Xnr3 at blastula confirming that Rac1 is required for early Wnt signaling. (I-I’) The colorectal cancer cell line SW480, in which the Wnt signaling pathway is activated by APC mutation, was positive for Rac1 and nuclear β-catenin immunostaining. (I’’) Merged panel including the DNA stain DAPI. (J-J’’) SW480 cells treated with the EIPA inhibitor (40 μM) had strongly decreased Rac1 and nuclear β-catenin levels, indicating a requirement for macropinocytosis. The numbers of embryos analyzed were as follows: A = 52, 100%; B = 47, 95% with ventralized small head phenotype; D = 58, 100%; E = 67; 97%; F = 64, 96%; G = 62, 97%; four independent experiments. Scale bars for embryos 500 μm; scale bars for immunofluorescence, 10 μm. Experiments represent biological replicates. Error bars denote SEM (n ≥ 3) (**P < 0.01). Also see Figure S2.

Microinjection of a small amount of EHT (1 mM, 4 nl, 1x ventral at 4-cell) was without effect on its own but was able to block the axis-duplication effect of microinjecting Wnt8 mRNA (Figure 4D-G). Quantitative RT-PCR analyses at the blastula stage confirmed that inhibiting Rac1 inhibited the induction of Siamois and Xnr3 caused by xWnt8 mRNA microinjection (Figure 4H). In SW480 CRC cells, which have constitutively activated Wnt due to mutation of APC (Leibovitz et al., 1976; Faux et al., 2004), EHT inhibited the size of cell spheroids in hanging drop cultures (Figure S2A-C) and, importantly, significantly inhibited macropinocytosis uptake of TMR-dextran in these cancer cells (Figure S2D-F). SW480 cells have high constitutive nuclear β-catenin and Rac1 plasma membrane levels (Figure 4I-I’’). After overnight treatment with the macropinocytosis inhibitor EIPA, both proteins decreased greatly, indicating a requirement of sustained macropinocytosis for the stability of these proteins (Figure 4J-J’’).

The results show that signaling by Wnt/β-catenin requires macropinocytosis, Rac1, and membrane trafficking in Xenopus embryos and in cancer cells with constitutive Wnt signaling.

Tumor Progression in Human Colorectal Cancers Correlates with a Role for membrane Trafficking in Wnt/β-catenin Signaling

Colorectal cancer is particularly favorable for the analysis of tumor progression because it gradually accumulates mutations, and in over 85% of the cases the initial driver mutation is in the APC tumor suppressor (Kinzler and Volgelstein, 1996; Weinberg, 2007). This offered an opportunity to test our proposal that membrane trafficking and its molecular components play an important role in Wnt-driven cancers (Albrecht et al., 2021; Tejeda-Muñoz and De Robertis, 2022b). Commercially available arrays of paraffin-sectioned histological sections containing 90 cases of adenocarcinoma of various grades I-IV and 90 samples of corresponding adjacent normal tissues (from TissueArray) were used for immunostaining. Arrays were double stained for β-catenin and antibodies against proteins involved in the macropinocytosis/MVB/lysosome/GSK3 pathway. Each individual section was evaluated, and normal colon, adenocarcinoma I, and adenocarcinoma IV were compared to assess the predominant effect of cancer progression.

As shown in Figure 5, β-catenin levels consistently correlated with grade IV malignancy when compared to normal colon or grade I cancers. Pak1, a kinase required for macropinocytosis and Wnt signaling (Redelman-Sidi et al., 2018; Albrecht et al.,2020) was significantly elevated with increased malignancy (Figure 5A-C’’). CD63, a tetraspan protein that marks MVB intraluminal vesicles (Escola et al., 1998), reached high levels in adenocarcinoma IV (Figure 5D-F’’), particularly in cells with the highest β-catenin expression (compare Figure 5F to 5F’). This increase in MVBs is in agreement with the GSK3 sequestration model of Wnt signaling (Taelman et al., 2010). The V-ATPase subunit V0a3, a marker for acidic lysosomes (Ramirez et al., 2019), also increased strongly with the CRC malignancy (compare Figure 5G to 5I). The cellular levels of total GSK3 decreased with malignancy (Figure 5J-L’’, compare 5J to 5L), as predicted by the sequestration model (Taelman et al., 2010). During Wnt signaling, GSK3 is sequestered into MVBs which are trafficked into lysosomes and a decrease was to be expected after sustained Wnt activation. Previously, we had been unable to demonstrate a decrease in total cellular GSK3 levels during acute Wnt signaling experiments (Taelman et al., 2010). The co-localization of β-catenin with Pak1, CD63, V0a3, or GSK3 in human colorectal adenocarcinomas was quantified in Figures S3A-S3D.

Degree of Human CRC Malignancy is Associated with Increased Macropinocytosis/MVB/lysosome markers and decreased GSK3 levels (A-A’) Normal human colon paraffin section stained with the macropinocytosis marker Pak1 and β-catenin, respectively. (A’’) Merged image with DAPI, a few cells colocalize both markers. (B-B’’) Pak1 and β-catenin levels are moderately increased at an early stage I CRC adenocarcinoma. (C’-C’’) Strong colocalization between Pak1 and β-catenin was observed in advanced stage IV CRC (inset). (D-D’’) Normal human colon section stained with the MVB marker CD63 and β-catenin; colocalization was not observed. (E-E’’) CD63 and β-catenin were stabilized in adenocarcinoma I, and moderate colocalization was found between CD63 and β-catenin. (F-F’) CD63 and β-catenin were strongly stabilized and colocalized in adenocarcinoma stage IV CRC. (F’’) Merge; note the striking colocalization between the MVB marker CD63 and β-Catenin in advanced stages of cancer (inset). (G-G’’) Normal colon stained for V0a3 (a subunit of V-ATPase that marks lysosomes) and β-catenin. (H-H’’) Stage I adenocarcinoma with moderately increased levels of lysosomes and β-catenin. (I-I’’) Strong co-localization of lysosomes and β-catenin in stage IV CRC (see inset). (J-J’’) Human colon array stained with the GSK3 and β-catenin. (K-K’’) GSK3 decreases, and β-catenin increases, at early stages of carcinogenesis, (L-L’’) GSK3 levels are very low in advanced CRC compared to normal human colon, while β-catenin levels are very high. Scale bars, 10 μm. Also see Figure S3 for quantifications and mouse xenografts of CRC cells.

These results were confirmed by examining immunostaining levels of Pak1, V0a3, and GSK3 compared to β-catenin protein in mouse xenografts in a CD1 NU/NU nude mouse model injected with SW480 human CRC cells (in which Wnt activation is known to be caused by APC mutation, Faux et al., 2004). In this case, tumor cells were compared to normal colon mouse tissue sections. The results showed that CRC cells had high levels of β-catenin, Pak1, and lysosomes, while GSK3 levels were reduced (Figure S3E-M), in agreement with the results in human tissue arrays.

These in vivo observations in human tumors are consistent with the view that macropinocytosis components, MVBs, and lysosomes increase with CRC malignancy. Further, the decrease in total GSK3 levels suggests a role in the trafficking and degradation of this key enzyme in cancer.

Wnt Affects the Subcellular Localization of Focal Adhesion Components

Previous work from our lab showed that Wnt treatment rapidly induces the endocytosis of multiple focal adhesion (FA) proteins, including integrin β-1 (ITGβ-1), from the cell surface (Tejeda-Muñoz et al., 2022). FAs link the actin cytoskeleton with the extracellular matrix (Figure 6A). Some focal adhesion proteins shuttle between the nucleus, cytoplasm and focal adhesion sites, and are proposed to play a role in cancer (Nix and Beckerle, 1997). FA components containing LIM domains can shuttle between the nucleus and cytoplasm (Kadrmas and Beckerle, 2004; Anderson et al., 2021). We now investigated whether constitutive Wnt signaling regulates FAs trafficking using the SW480 and SW480APC system, in which the stable reconstitution with full-length APC restores these CRC cells to a non-malignant phenotype (Faux et al., 2004), which is accompanied by a decrease in β-catenin levels (Figure 6B). As shown in Figure 6C to 6D’’, the PET-LIM domain FA protein Tes (also known as Testin) (Garvalov et al., 2003) was cytoplasmic in SW480APC, but was nuclear in SW480 cells in which Wnt is activated by APC mutation. The nuclear localization of Tes in SW480 cells required macropinocytosis, since treatment with EIPA resulted in its re-localization to the cytoplasm (Figure 4E-4F’’). This result was consistent with the idea that membrane trafficking is required for constitutive Wnt signaling in cancer (Tejeda-Muñoz et al., 2019). In the nude mouse xenograft model, Tes was abundant in the cytoplasm of the normal colon, while in SW480 it localized in the nucleus of those cells with the highest β-catenin levels (Figure 6G-6I’’). The function of TES is poorly understood, but it has been proposed to act as a tumor suppressor (Tatarelli et al., 2000; Tobias et al., 2001); the present results suggest that Tes may serve as a Wnt-regulated nuclear marker in CRC.

A Focal Adhesion Protein Changes Its Nucleocytoplasmic Distribution after Wnt Signaling Activation, and FAK is Stabilized by PMA (A) Diagram showing how focal adhesions connect the actin cytoskeleton to the extracellular matrix via integrins; Wnt signaling changes the distribution of the focal adhesion protein Tes from focal adhesion sites to the nucleus. (B) Western blot of SW480 cells with and without stable expression of APC; note the reduction in Tes expression and increase in β-catenin levels in SW480 cells lacking full-length APC. GADPH was used as loading control. (C-C’) Colon cancer cell line SW480 stably restored with full-length APC stained with Tes and β-catenin antibodies; note that in SW480APC cells TES is absent from the nucleus (inset). (D-D’’) In SW480 CRC cells both Tes and β-catenin are located the nucleus. (E-E’’) Colon cancer cell line SW480 stained with Tes (in green) and β-catenin (in red) antibodies as a control. (F-F’’) Inhibiting macropinocytosis with EIPA (40 μM) treatment restored the cytoplasmic localization of the focal adhesion protein Tes in SW480 cells and strongly inhibited β-catenin levels. (G) Quantification of colocalization between Tes and β-catenin in normal mouse colon and SW480 xenografts. (H-H’) Immunohistochemistry of the CD1 NU/NU normal mouse colon showing Tes and β-catenin distribution. (H’’) Merge showing modest co-localization (inset). (I-I’’) Immunohistochemistry images of the CD1 NU/NU mouse xenograft model showing nuclear TES, which is strongly colocalized with β-catenin levels in cancer cells (inset). (J) Reporter assay in HEK293T cells stably expressing BAR and Renilla, showing strong inhibition of the β-catenin transcriptional activity induced by LiCl (40 mM) by FAK inhibitor PF-00562271 (20 μM) after overnight treatment. See also Figure S4A-S4B’’. (K-K’’) SW480 cells immunostained with FAK and β-catenin antibodies. (L-L’’) The tumor promoter PMA (750 nM, overnight treatment) increased levels of both β-catenin and FAK in SW480 cells (in which the Wnt pathway is constitutively activated by APC mutation). This increase in FAK levels requires membrane macropinocytosis and lysosome activity (see Figure S4C-S4H’’). Error bars denote SEM (n ≥ 3) (**P < 0.01) (Scale bars, 10 μm). Also see Figure S4.

Focal Adhesion Kinase (FAK) is a key regulator of FA maintenance and signaling, known to interact with the Wnt pathway in various ways (Chuang et al., 2022). In HEK293 BAR/Renilla cells treatment with FAK inhibitor PF-00562271 strongly decreased β-catenin signaling by LiCl inhibition of GSK3 (Figure 6J). In SW480 cells, FAK had nuclear and cytoplasmic localization (Figure 6A-A’’). Overnight treatment with FAK inhibitor PF-00562271 resulted in the reduction of β-catenin and of FAK itself (Figure S4B-B’’). Importantly, the addition of the tumor promoter PMA to SW480 cells (in which the Wnt pathway is constitutively activated) strongly increased protein levels of β-catenin and FAK protein (Figure 6K-6L’’). The effect of PMA was abrogated by EIPA or Baf treatment (Figure S4C-S4H’’), indicating that the increase in FAK levels requires macropinocytosis and lysosome activity.

The shuttling of FA components between cytoplasm and nucleus when Wnt signaling is activated by APC mutation, and increased by the macropinocytosis-inducing tumor promoter PMA, is probably just the tip of the iceberg of the role of membrane trafficking in Wnt-driven cancers.

Discussion

The present work was inspired by the discovery that PMA increases macropinocytosis (Swanson, 1989). Our experiments support the role of membrane trafficking in canonical Wnt signaling in embryos and cancer cells and are summarized in the model in Figure 7. The main finding was a synergistic effect between the tumor promoter phorbol ester (PMA) and activation of the Wnt/β-catenin pathway that was blocked by EIPA or Baf (Figure 1). EIPA is a macropinocytosis inhibitor of the Na+/H+ exchanger at the plasma membrane; it causes acidification of the submembranous cytoplasm and prevents actin polymerization (Koivusalo et al., 2010). Bafilomycin A (Baf) is an inhibitor of the V-ATPase that drives membrane trafficking and lysosomal acidification. PMA plays an important role in our understanding of tumor biology because it is the archetypal tumor promoter, able to promote tumor growth without further mutation of the DNA (Berenblum, 1941; Weinberg, 2007). PMA activates the PKC pathway (Takai et al., 1979; Nishizuka, 1984), and there is previous literature on the crosstalk between the PKC and Wnt pathways (Schwarz et al., 2013; Goode et al., 1992). What is novel in the present paper is the connection between macropinocytosis, membrane trafficking, lysosomes, focal adhesions, PMA, and Wnt signaling.

Model of PMA Synergy with Wnt Signaling through Macropinocytosis and Membrane Trafficking Wnt and PMA are activators of macropinocytosis, which is a cell drinking process driven by an actin meshwork that requires the small GTPase Rac1. Membrane trafficking into MVBs/lysosomes (marked by CD63) requires acidification by V-ATPase, which is inhibited by Bafilomycin A (Baf). Phorbol esters such as PMA stimulate macropinocytosis and synergize with Wnt signaling. Inhibiting macropinocytosis with the membrane trafficking inhibitors EIPA, Baf, or the Rac1 inhibitor EHT1864 block Wnt signaling and its cooperation with the tumor promoter PMA.

Macropinocytosis, Rac1, and Wnt signaling

Wnt signaling triggers lysosomal acidification in addition to macropinocytosis (Albrecht et al., 2020). Baf blocks the synergy between PMA and the Wnt-mimic LiCl in embryos microinjections.

It also blocks development of the endogenous axis, particularly the head region when embryos are briefly immersed in Baf at the 32-cell stage (Tejeda-Muñoz and De Robertis, 2022). Using order-of-addition experiments in which embryos were briefly treated with pulses of Baf or the GSK3 inhibitor LiCl, we showed that the requirement for lysosome acidification was due to inhibition of the Wnt pathway at the 32-cell stage (Figure 2). Co-injection experiments of DN-GSK3 or β-catenin mRNAs with DN-Rab7 or EIPA supported the view that membrane trafficking and macropinocytosis are essential for the induction of secondary axes in Xenopus (Figure 3).

Rac1 is an upstream regulator of the Pak1 kinase required for macropinocytosis (Redelman-Sidi et al., 2018). Rac1 is a member of the family of small Rho GTPases that drives the formation of lamellipodia, membrane ruffles, and macropinocytosis (Hall, 1998; Egami et al., 2014; Fujii et al., 2013). The Rac1 inhibitor EHT1864 (Hampsch et al., 2017) strongly ventralized the endogenous axis of embryos after brief incubation at the critical 32-cell stage. In co-injection experiments EHT blocked the axis-inducing effect of microinjected xWnt8 mRNA (Figure 4). Rac1 is a regulator of multiple signaling cascades and has been reported to increase the nuclear translocation of β-catenin (Wu et al., 2008). Other work showed that Rac1 binds to β-catenin and promotes the formation of nuclear β-catenin/LEF1 complexes (Jamieson et al., 2015). In the context of the present study, Rac1 levels were decreased by treating CRC cells with EIPA, and Rac1 inhibition blocked Wnt signaling in embryos, indicating that Rac1 may crosstalk with Wnt through macropinocytosis.

Lysosomes, cancer and focal adhesions

The lysosome is considered a promising target for the treatment of cancer (Fehrenbacher and Jäättelä, 2005; Lawrence and Zoncu, 2019). We used human CRC tissue arrays to analyze whether lysosomes and membrane trafficking correlate with cancer progression. Most colorectal tumors are initiated by mutation of APC, which drives Wnt signaling (Kinzler and Vogelstein, 1996; Segditsas and Tomlinson, 2006). It was found that advanced grades of colon carcinoma correlated with increase immunostaining for Pak1 (driver of macropinocytosis), CD63 (MVB/late endosome marker), and V0a3 (lysosomal V-ATPase marker) (Figure 5). In addition, cancer malignancy correlated with low levels of GSK3 staining, both in human samples and nude mouse SW480 xenografts. Previous literature points to the physiological relevance of lysosomes during cancer. Cancer cells require lysosome function and have changes in lysosomal volume and subcellular localization during oncogenic transformation (Kallunki et al., 2012; Kirkegaard and Jäättelä, 2009; Gocheva and Joyce, 2007; Kolwijck et al., 2010). Lysosome-associated membrane protein 1 (LAMP-1) has been reported on the cell surface of highly metastatic colon cells (Ranjbar and Jamshidi, 2022) and V-ATPase has been associated with extracellular acidification and cancer cell invasion (Capecci and Forgac, 2013; Hinton et al., 2009; Perona and Serrano, 1988). The results reported here support the view (Taelman et al., 2010; Tejeda-Muñoz and De Robertis, 2022) that Wnt-driven lysosomal acidification and GSK3 sequestration may play a role in cancer. In addition, MVB/lysosome components could serve as markers for advanced Wnt-driven cancers.

The cytoskeleton interacts with proteins of the extracellular matrix though focal adhesions (Geiger et al., 2001). We have previously reported that Wnt causes the endocytosis of FA proteins and depletion of Integrin β-1 from the cell surface (Tejeda-Muñoz et al., 2022). We now report that Tes, a member of the PET-LIM family of focal adhesion proteins, changed subcellular localization after Wnt activation in SW480 cells. It is known that other focal adhesions shuttle between the nucleus, cytoplasm, and focal adhesion sites and may play important roles in cancer (Chaturvedi et al., 2012; Nix and Beckerle, 1997). FAK is a key signaling kinase that regulates focal adhesion signaling. In SW480 cells, which have constitutive Wnt signaling, FAK protein was increased by PMA, and this was inhibited by EIPA or Baf (Figure 6). These results suggest that the crosstalk between focal adhesions and Wnt signaling could be relevant in cancer progression.

Cooperation between PMA and Wnt

PMA activates PKC signaling by mimicking its physiological second messenger diacylglycerol (DAG), an activator of protein kinase C. There is previous literature implicating PKC isozymes in the regulation of the Wnt signaling pathway (Fang et al., 2002; Goode et al., 1992) and that different PKC isoforms can have positive or negative effects on Wnt/β-catenin signaling (Ohno and Nishizuka, 2002). However, the molecular role of PKC isoforms in the Wnt signaling pathway remains poorly understood (Bhatia and Spiegelman, 2005; Schwarz et al., 2013) It has been reported that PMA can stabilize CK1ε (Casein Kinase 1 epsilon), enhance its kinase activity, and induce LRP6 phosphorylation at Thr1479 and Ser1490, leading to the activation of the Wnt/β-catenin pathway (Su et al., 2018).

In the context of the present study, it is relevant that Wnt signaling rapidly triggers macropinocytosis and trafficking into lysosomes (Tejeda-Muñoz et al., 2019). Macropinocytosis requires the activation of Phosphoinositide 3 Kinase (PI3K) which leads to the formation of patches of PIP3 in the inner plasma membrane leaflet that initiate the formation of macropinocytic cups (Egami et al., 2014; Yoshida et al., 2018). During macropinosome formation, the PIP3 lipid is then converted into DAG by Phospholipase C gamma 1 (PLCγ1), and DAG activates PKC signaling (Yoshida et al., 2018). It seems possible that PMA may facilitate Wnt/β-catenin signaling by mimicking the DAG concentrated in macropinosomes. A future test for this hypothesis will be to directly add DAG to cells and follow its effects on Wnt/β-catenin signaling. In addition, it will be interesting to study the role of chronic inflammation on macropinocytosis and tumor promotion. Active PKC phosphorylates many targets, including IκB Kinase (IKK) which becomes activated and phosphorylates Inhibitor of NF-κB (IκB), targeting it for degradation and triggering inflammation (Weinberg, 2007). Tumor progression is promoted by many chronic inflammatory agents such as turpentine, distilled alcoholic beverages, hepatitis viruses, and Helicobacter pylori that do not cause mutations by themselves but regulate the expansion of previously mutant clones (Rous and Kidd, 1941; Weinberg, 2007). Taken together, the results presented here suggest that macropinocytosis, lysosomal activity, and membrane trafficking are possible therapeutic targets for tumor progression in Wnt-driven cancers.

eLife Assessment

This valuable study describes an interesting synergism between macropinocytosis and Wnt signaling in multiple biological systems. The main claims are at least partially supported with solid evidence. The pharmacological manipulations comes with a number of caveats, and the mechanistic basis of the described synergism remains unclear. The study will be of interest to cell biologists and biomedical researchers, particularly in the Wnt field and in tumor biology.

Reviewer #1 (Public Review):

In this ms, Tejeda-Muñoz and colleagues examine the roles of macropinocytosis in WNT signalling activation in development (Xenopus) and cancer (CRC sections, cell lines and xenograft experiments). Furthermore, they investigate the effect of the inflammation inducer Phorbol-12-myristate-13-acetate (PMA) in WNT signalling activation through macropinocytosis. They propose that macropinocytosis is a key driver of WNT signalling, including upon oncogenic activation, with relevance in cancer progression.

I found the analyses and conclusions of the relevance of macropinocytosis in WNT signalling compelling, notably upon constitutive activation both during development and in CRC. However, I think this manuscript only partially characterises the effects of PMA in WNT signalling, largely due to a lack of an epistatic characterisation of PMA roles in Wnt activation. For example:

1- The authors show that PMA cooperate with 1) GSK3 inhibition in Xenopus to promote WNT activation, and 2) (possibly) with APCmut in SW480 to induce b-cat and FAK accumulation. To sustain a specific functional interaction between WNT and PMA, the effects should be tested through additional epistatic experiments. For example, does PMA cooperate with Wnt8 in axis duplication analyses? Does PMA cooperate with any other WNT alteration in CRC or other cell lines? Importantly, does APC re-introduction in SW480 rescue the effect of PMA? Such analyses could be critical to determine specificity of the functional interactions between WNT and PMA. This question could be addressed by performing classical epistatic analyses in cell lines (CRC or HEK) focusing on WNT activity, and by including rescue experiments targeting the WNT pathway downstream of the effects e.g., dnTCF, APC re- introduction, etc.

2- While the epistatic analyses of WNT and macropinocytosis are clear in frog, the causal link in CRC cells is contained to b-catenin accumulation. While is clear that macropinocytosis reduces spheroid growth in SW480, the lack of rescue experiments with e.g., constitutive active b-catenin or any other WNT perturbation or/and APC re-introduction, limit the conclusions of this experiment.

Minor comments:

3- Different compounds targeting membrane trafficking are used to rescue modes of WNT activation (Wnt8 vs LiCl) in Xenopus.

4- The abstract does not state the results in CRC/xenografts

5- Labels of Figure 2E might be swap

6- Figure 4i,j, 6 and s4 rely on qualitative analyses instead of quantifications, which underscores their evaluation. On the other hand, the detailed quantifications in Figure S3A-D strongly support the images of Figure 5

Reviewer #2 (Public Review):

Tejeda Muñoz et al. investigate the intersection of Wnt signaling, macropinocytosis, lysosomes, focal adhesions and membrane trafficking in embryogenesis and cancer. Following up on their previous papers, the authors present evidence that PMA enhances Wnt signaling and embryonic patterning through macropinocytosis. Proteins that are associated with the endo-lysosomal pathway and Wnt signaling are co-increased in colorectal cancer samples, consistent with their pro-tumorigenic action. The function of macropinocytosis is not well understood in most physiological contexts, and its role in Wnt signaling is intriguing. The authors use a wide range of models - Xenopus embryos, cancer cells in culture and in xenografts and patient samples to investigate several endolysosomal processes that appear to act upstream or downstream of Wnt. A downside of this broad approach is a lack of mechanistic depth. In particular, few experiments monitor macropinocytosis directly, and macropinocytosis manipulations have pleiotropic effects that are open alternative interpretations. Several experiments are confirmatory of previous findings; the manuscript could be improved by focusing on the novel relationship between PMA-induced macropinocytosis and better support these conclusions with additional experiments.

The authors use a range of inhibitors that suppress macropinosome formation (EIPA, Bafilomycin A1, Rac1 inhibition). However, these are not specific macropinocytosis inhibitors (EIPA blocks an Na+/H+ exchanger, which is highly toxic and perturbs cellular pH balance; Bafilomycin blocks the V-ATPase, which has essential functions in the Golgi, endosomes and lysosomes; Rac1 signals through multiple downstream pathways). A specific macropinocytosis inhibitor does not exist, and it is thus important to support key conclusions with dextran uptake experiments.

The title states that PMA increases Wnt signaling through macropinocytosis. However, the mechanistic relationship between PMA-induced macropinocytosis and Wnt signaling is not well supported. The authors refer to a classical paper that demonstrates macropinocytosis induction by PMA in macrophages (PMID: 2613767). Unlike most cell types, macrophages display growth factor-induced and constitutive macropinocytic pathways (PMID: 30967001). It would thus be important to demonstrate macropinocytosis induction by PMA experimentally in Xenopus embryos / cancer cells. Does treatment with EIPA / Bafilomycin / Rac1i decrease the dextran signal in embryos? In macrophages, the PKC inhibitor Calphostin C blocks macropinocytosis induction by PMA (PMID: 25688212). Does Calphostin C block macropinocytosis in embryos / cancer cells? Do the various combinations of Wnts / Wnt agonists and PMA have additive or synergistic effects on dextran uptake? If the authors want to conclude that PMA activates Wnt signaling, it would also be important to demonstrate the effect of PMA on Wnt target gene expression.

The experiments concerning macropinosome formation in Xenopus embryos are not very convincing. Macropinosomes are circular vesicles whose size in mammalian cells ranges from 0.2 - 10 µM (PMID: 18612320). The TMR-dextran signal in Fig. 1A does not obviously label structures that look like macropinosomes; rather the signal is diffusely localized throughout the dorsal compartment, which could be extracellular (or perhaps cytosolic). I have similar concerns for the cell culture experiments, where dextran uptake is only shown for SW480 spheroids in Fig. S2. It would be helpful to quantify size of the circular structures (is this consistent with macropinosomes?).

In Fig. 4I - J, the dramatic decrease in b-catenin and especially in Rac1 after overnight EIPA treatment is rather surprising. How do the authors explain these findings? Is there any evidence that macropinocytosis stabilizes Rac1? Could this be another effect of EIPA or general toxicity?

On a similar note, Fig. 6 K - L the FAK staining in control cells appears to localize to focal adhesions, but in PMA-treated cells is strongly localized throughout the cell. Do the authors have any thoughts on how PMA stabilizes FAK and where the kinase localizes under these conditions? Does PMA treatment increase FAK signaling activity?

The tumor stainings in Figure 5 are interesting but correlative. Pak1 functions in multiple cellular processes and Pak1 levels are not a direct marker for macropinocytosis. In the discussion, the authors discuss evidence that the V-ATPase translocates to the plasma membrane in cancer to drive extracellular acidification. To which extent does the Voa3 staining reflect lysosomal V-ATPase? Do the authors have controls for antibody specificity?

Reviewer #3 (Public Review):

The manuscript by Tejeda-Munoz examines signaling by Wnt and macropinocytosis in Xenopus embryos and colon cancer cells. A major problem with the study is the extensive use of pleiotropic inhibitors as "specific" inhibitors of macropinocytosis in embryos. It is true that BafA and EIPA block macropinocytosis, but they do many other things as well. A major target of EIPA is the NheI Na+/proton transporter, which also regulates invasive structures (podosomes, invadopodia) which could have major roles in development. Similarly, Baf1 will disrupt lysosomes and the endocytic system, which secondary effects on mTOR signaling and growth factor receptor trafficking. The authors cannot assume that processes inhibited by these drugs demonstrate a role of macropinocytosis. While correlations in tumor samples between increased expression of PAK1 and V0a3 and decreased expression of GSK3 are consistent with a link between macropinocytosis and Wnt-driven malignancy, the cell and embryo-based experiments do not convincingly make this connection. Finally, the data on FAK and TES are not well integrated with the rest of the manuscript.

1. The data in Fig. 1A do not convincingly demonstrate macropinocytosis - it is impossible to tell what is being labeled by the dextran.

2. The data in Fig. 2 do not make sense. LiCL2 bypasses the WNT activation pathway by inhibiting GSK3. If subsequent treatment with BafA blocks the effects of GSK3 inhibition, then BafrA is doing something unrelated to Wnt activation, whose target is the inhibition/sequestration of GSK3. While BafA might block GSK3 sequestration by inhibiting MVB function, it should have no effect on the inhibition of GSK3 by LiCl2.

3. The effect of EHT on MP in SW480 cells is not clearly related to what is happening in the embryos. The nearly total loss of staining for Rac and -catenin after overnight EIPA does not implicate MP in protein stability - critical controls for cell viability and overall protein turnover are absent. Inhibition of WNT signaling might be expected to enhance -catenin turnover, but the effect on Rac1 is surprising. A more quantitative analysis by western blotting is required.

4. The data on FAK inhibition and TES trafficking are poorly integrated with the rest of the paper.

Author Response

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

In this ms, Tejeda-Muñoz and colleagues examine the roles of macropinocytosis in WNT signalling activation in development (Xenopus) and cancer (CRC sections, cell lines and xenograft experiments). Furthermore, they investigate the effect of the inflammation inducer Phorbol-12-myristate-13-acetate (PMA) in WNT signalling activation through macropinocytosis. They propose that macropinocytosis is a key driver of WNT signalling, including upon oncogenic activation, with relevance in cancer progression.

I found the analyses and conclusions of the relevance of macropinocytosis in WNT signalling compelling, notably upon constitutive activation both during development and in CRC.

Thank you.

However, I think this manuscript only partially characterises the effects of PMA in WNT signalling, largely due to a lack of an epistatic characterisation of PMA roles in Wnt activation. For example: 1- The authors show that PMA cooperate with 1) GSK3 inhibition in Xenopus to promote WNT activation, and 2) (possibly) with APCmut in SW480 to induce b-cat and FAK accumulation. To sustain a specific functional interaction between WNT and PMA, the effects should be tested through additional epistatic experiments. For example, does PMA cooperate with Wnt8 in axis duplication analyses? Does PMA cooperate with any other WNT alteration in CRC or other cell lines? Importantly, does APC re-introduction in SW480 rescue the effect of PMA? Such analyses could be critical to determine specificity of the functional interactions between WNT and PMA. This question could be addressed by performing classical epistatic analyses in cell lines (CRC or HEK) focusing on WNT activity, and by including rescue experiments targeting the WNT pathway downstream of the effects e.g., dnTCF, APC re- introduction, etc.

We agree that there was need for additional direct evidence of functional interactions of between macropinocytosis, Wnt signaling, and PMA beyond the previously provided target gene assays in Xenopus (now shown in Figure 1I) and luciferase assays in cultured cells (Figure 1J) which used LiCl and inhibition by Bafilomycin. We therefore carried out a new experiment using 3T3 cells, now shown in Figure 1K-P. Wnt3a protein increased the uptake of TMR-dextran 70 kDa, and PMA enhanced this response. The macropinocytosis inhibitor EIPA blocked induction of macropinocytosis by Wnt3a and PMA. These results were quantitated in Figure 1Q. We think this new experiment strengthens the main conclusion that the tumor promoter PMA increases macropinocytosis. Thank you.

1. While the epistatic analyses of WNT and macropinocytosis are clear in frog, the causal link in CRC cells is contained to b-catenin accumulation. While is clear that macropinocytosis reduces spheroid growth in SW480, the lack of rescue experiments with e.g., constitutive active b-catenin or any other WNT perturbation or/and APC re-introduction, limit the conclusions of this experiment.

We now provide new experiments in 3T3 cells treated with LiCl, overexpression of constitutively-active β-catenin and constitutively-active Lrp6 (Figure 4, panels I through L’’); the new results indicate that Wnt signaling activation increases protein levels of the macropinocytosis activator Rac1.

Minor comments:

3- Different compounds targeting membrane trafficking are used to rescue modes of WNT activation (Wnt8 vs LiCl) in Xenopus.

The main goal of our experiments was to test the requirement of membrane trafficking for tumor promoter activity through the Wnt pathway. We therefore used PMA, and a variety of inhibitors such as EIPA (Na+/H+ exchanger, Figure 1I and Figure 3D), Bafilomycin A (Figure 1H), DN-Rab7 (Figure 3G) and EHT1864 (a Rac1 inhibitor, Figure 4G). One could argue that using a wide variety of membrane trafficking inhibitors is a plus.

4- The abstract does not state the results in CRC/xenografts

We have added a sentence to the abstract.

5- Labels of Figure 2E might be swap

Thank you for detecting this error, we now label the last two columns in Figure 2E correctly.

6- Figure 4i,j, 6 and s4 rely on qualitative analyses instead of quantifications, which underscores their evaluation. On the other hand, the detailed quantifications in Figure S3A-D strongly support the images of Figure 5

The quantifications of the previous Figure 4I-J supported the data in the initial reviewed preprint, shown in Author response image 1:

However, these data have now been deleted from this version to make space for new experiments showing the stabilization of Rac1 by stabilized β-catenin and CA-LRP6. Quantifications in Figure 6C-F’’ are not shown because they represent changes in subcellular localization, but a western blot is provided in Figure 6B. Quantifications for Figure 6H-I’’ are shown in panel 6G. Supplemental Figure S4 already has 24 panels so introducing quantifications would be unwieldy.

Thank you for the thoughtful comments.

Reviewer #2 (Public Review):

Tejeda Muñoz et al. investigate the intersection of Wnt signaling, macropinocytosis, lysosomes, focal adhesions and membrane trafficking in embryogenesis and cancer. Following up on their previous papers, the authors present evidence that PMA enhances Wnt signaling and embryonic patterning through macropinocytosis. Proteins that are associated with the endo-lysosomal pathway and Wnt signaling are co-increased in colorectal cancer samples, consistent with their pro-tumorigenic action. The function of macropinocytosis is not well understood in most physiological contexts, and its role in Wnt signaling is intriguing. The authors use a wide range of models - Xenopus embryos, cancer cells in culture and in xenografts and patient samples to investigate several endolysosomal processes that appear to act upstream or downstream of Wnt. A downside of this broad approach is a lack of mechanistic depth. In particular, few experiments monitor macropinocytosis directly, and macropinocytosis manipulations have pleiotropic effects that are open alternative interpretations. Several experiments are confirmatory of previous findings; the manuscript could be improved by focusing on the novel relationship between PMA-induced macropinocytosis and better support these conclusions with additional experiments.

New additional experiments focusing on the role of PMA are now provided.

The authors use a range of inhibitors that suppress macropinosome formation (EIPA, Bafilomycin A1, Rac1 inhibition). However, these are not specific macropinocytosis inhibitors (EIPA blocks an Na+/H+ exchanger, which is highly toxic and perturbs cellular pH balance; Bafilomycin blocks the V-ATPase, which has essential functions in the Golgi, endosomes and lysosomes; Rac1 signals through multiple downstream pathways). A specific macropinocytosis inhibitor does not exist, and it is thus important to support key conclusions with dextran uptake experiments.

We used a wide range of inhibitors because the main idea is to show that membrane trafficking is important in Wnt and PMA activity. We would like to point out that the current experimental definition in the field of macropinocytosis, despite any caveats, is the ability to block dextran uptake with EIPA. Because inhibitors may not be entirely specific, we think using a broad approach to target membrane trafficking might be a plus. We now provide in Figure 1K-Q a new experiment showing that Wnt3a protein treatment increases dextran uptake and PMA stimulates this macropinocytosis in 3T3 cells. EIPA inhibited dextran macropinocytosis in the presence of Wnt and PMA (Figure 1N and 1Q). We also provide a time-lapse video of the rapid macropinocytic vesicles induction by PMA in SW480 CRC cells in which the plasma membrane is tagged (Supplemental Movie S1).

The title states that PMA increases Wnt signaling through macropinocytosis. However, the mechanistic relationship between PMA-induced macropinocytosis and Wnt signaling is not well supported. The authors refer to a classical paper that demonstrates macropinocytosis induction by PMA in macrophages (PMID: 2613767). Unlike most cell types, macrophages display growth factor-induced and constitutive macropinocytic pathways (PMID: 30967001). It would thus be important to demonstrate macropinocytosis induction by PMA experimentally in Xenopus embryos / cancer cells. Does treatment with EIPA / Bafilomycin / Rac1i decrease the dextran signal in embryos? In macrophages, the PKC inhibitor Calphostin C blocks macropinocytosis induction by PMA (PMID: 25688212). Does Calphostin C block macropinocytosis in embryos / cancer cells? Do the various combinations of Wnts / Wnt agonists and PMA have additive or synergistic effects on dextran uptake? If the authors want to conclude that PMA activates Wnt signaling, it would also be important to demonstrate the effect of PMA on Wnt target gene expression.

We now provide a new experiment showing macropinocytosis induction of PMA experimentally in cancer cells. CRC SW480 cells, despite having a mutant APC, are able to respond to PMA by further increasing TMR-dextran 70 kDa uptake over background within 1 hour (now shown in Figure S1):

Investigating PKC and Calphostin C is outside of goals of this paper. With respect to final the point on the effect of PMA on Wnt target gene expression, this was shown in the context of the Xenopus embryo in Figure 1I (Siamois and Xnr3 are direct targets of Wnt).

The experiments concerning macropinosome formation in Xenopus embryos are not very convincing. Macropinosomes are circular vesicles whose size in mammalian cells ranges from 0.2 - 10 µM (PMID: 18612320). The TMR-dextran signal in Fig. 1A does not obviously label structures that look like macropinosomes; rather the signal is diffusely localized throughout the dorsal compartment, which could be extracellular (or perhaps cytosolic). I have similar concerns for the cell culture experiments, where dextran uptake is only shown for SW480 spheroids in Fig. S2. It would be helpful to quantify size of the circular structures (is this consistent with macropinosomes?).

In response, we have deleted the TMR experiments in Xenopus embryos; they will be reinvestigated at a later time. With respect to macropinosome sizes in cultured cells, they are indeed large at the plasma membrane level (see new Supplemental Movie S1), but rapidly decrease in size once dextran is concentrated inside the cell. This can be visualized in the new experiments showing dextran vesicles in Supplemental Figure S1J-K and Figure 1K-P.

In Fig. 4I - J, the dramatic decrease in b-catenin and especially in Rac1 after overnight EIPA treatment is rather surprising. How do the authors explain these findings? Is there any evidence that macropinocytosis stabilizes Rac1? Could this be another effect of EIPA or general toxicity?

We now provide new evidence that Wnt signaling stabilizes Rac1. The old data relying on overnight EIPA treatment has been replaced by new experiments in 3T3 cells showing (i) that LiCl treatment increases levels of Rac1 protein and β-catenin levels (Figure 4I-J’’), (ii) that cells transfected with constitutively active β-catenin-GFP have higher levels of Rac1 than control untransfected cells (Figure 4K-K’’) and (iii) that Rac1 is stabilized in cells transfected with CA-Lrp6-GFP when compared to untransfected cells (Figure4L-L’’).

On a similar note, Fig. 6 K - L the FAK staining in control cells appears to localize to focal adhesions, but in PMA-treated cells is strongly localized throughout the cell. Do the authors have any thoughts on how PMA stabilizes FAK and where the kinase localizes under these conditions? Does PMA treatment increase FAK signaling activity?

The previous Figure 6K-L’’ are now found in Supplementary Figure S4, panels C-D’’. The result is that FAK is greatly stabilized by overnight incubation with PMA. How this achieved is unknown, perhaps the result of increased macropinocytosis, but we do not wish to speculate in the main manuscript. We have not measured FAK activity, but the FAK inhibitor PF-00562271 strongly decreased β-catenin signaling by GSK3 inhibition (Figure 6J) and has strong effects in neural development that mimic inhibition of the early Wnt signal (new experiments shown in Figure 6K-L’’’). The results suggest that FAK activity affects Wnt signaling and dorsal development; the molecular mechanism of this interaction is unknown but worthy of future studies.

The tumor stainings in Figure 5 are interesting but correlative. Pak1 functions in multiple cellular processes and Pak1 levels are not a direct marker for macropinocytosis. In the discussion, the authors discuss evidence that the V-ATPase translocates to the plasma membrane in cancer to drive extracellular acidification. To which extent does the Voa3 staining reflect lysosomal V-ATPase? Do the authors have controls for antibody specificity?

It is true that Pak1 has multiple functions, yet it is essential for the actin machinery that drives macropinocytosis. We have now rephrased the discussion to say “Rac1 is an upstream regulator of the Pak1 kinase required for the actin machinery that drive macropinocytosis (Redelman-Sidi et al., 2018)”. We also explain that: “V-ATPase has been associated with acidification of the extracellular milieu in tumors (Capecci and Forgac, 2013; Hinton et al., 2009; Perona and Serrano, 1988). Extracellular acidification is probably due to increased numbers of lysosomes which are exocytosed, since V0a3 was located within the cytoplasm in advanced cancer or xenografts in mice (Figures 5I and S3I)”. The antibody we used for V0a3 is highly specific and has been used widely (Ramirez et al., 2019).

Reviewer #3 (Public Review):

The manuscript by Tejeda-Munoz examines signaling by Wnt and macropinocytosis in Xenopus embryos and colon cancer cells. A major problem with the study is the extensive use of pleiotropic inhibitors as "specific" inhibitors of macropinocytosis in embryos. It is true that BafA and EIPA block macropinocytosis, but they do many other things as well. A major target of EIPA is the NheI Na+/proton transporter, which also regulates invasive structures (podosomes, invadopodia) which could have major roles in development. Similarly, Baf1 will disrupt lysosomes and the endocytic system, which secondary effects on mTOR signaling and growth factor receptor trafficking. The authors cannot assume that processes inhibited by these drugs demonstrate a role of macropinocytosis. While correlations in tumor samples between increased expression of PAK1 and V0a3 and decreased expression of GSK3 are consistent with a link between macropinocytosis and Wnt-driven malignancy, the cell and embryo-based experiments do not convincingly make this connection. Finally, the data on FAK and TES are not well integrated with the rest of the manuscript.

The criticism that drugs are not entirely specific is a valid one. Our approach of using a variety of drugs such as EIPA, BafA, EHT1864 or FAK inhibitor PF-00562271 all point to the main conclusion that the membrane trafficking is important in signaling by Wnt and the action of the tumor promoter PMA. The data on FAK, TES and focal adhesions have been better integrated in the manuscript and new experiments on the effect of FAK inhibitor in embryonic dorsal development are now provided (Figure 6K-L’’’).

1. The data in Fig. 1A do not convincingly demonstrate macropinocytosis - it is impossible to tell what is being labeled by the dextran.

In response, we have deleted the TMR-dextran experiments in Xenopus embryos; they will be reported at a later time.

1. The data in Fig. 2 do not make sense. LiCL2 bypasses the WNT activation pathway by inhibiting GSK3. If subsequent treatment with BafA blocks the effects of GSK3 inhibition, then BafrA is doing something unrelated to Wnt activation, whose target is the inhibition/sequestration of GSK3. While BafA might block GSK3 sequestration by inhibiting MVB function, it should have no effect on the inhibition of GSK3 by LiCl2.

We now explain in the main text describing Figure 2 in the results, the initial effect of GSK3 inhibition by LiCl is to trigger macropinocytosis (Albrecht et al., 2020). If the downstream acidification of lysosomes is inhibited, then the brief treatment with LiCl (7 min at 32-cell stage) has no effect (LiCl 1st+BafA 2nd, Figure 2H). BafA inhibits lysosomal acidification at 32-cell stage resulting in ventralization, but the effect of brief BafA treatment can be reversed by inducing membrane trafficking by LiCl (BafA 1st+LiCl 2nd, Figure 2C). The labelling of the figure panels C and H has been modified to indicate this is an order-of-addition experiment. These order-of-addition experiments strongly support the proposal that endogenous lysosomal activity is required to generate the initial endogenous Wnt signal that takes place at the 32-cell stage of development (Tejeda-Muñoz and De Robertis, 2022a).

1. The effect of EHT on MP in SW480 cells is not clearly related to what is happening in the embryos. The nearly total loss of staining for Rac and -catenin after overnight EIPA does not implicate MP in protein stability - critical controls for cell viability and overall protein turnover are absent. Inhibition of WNT signaling might be expected to enhance -catenin turnover, but the effect on Rac1 is surprising. A more quantitative analysis by western blotting is required.

The results from SW480 cells inhibition by EIPA have been replaced in Figure 4. We now provide new evidence in 3T3 cells that Wnt signaling stabilizes Rac1. The old data relying on EIPA treatment in SW480 cells has been replaced by new experiments in 3T3 cells showing (i) that LiCl treatment increases levels of Rac1 protein and β-catenin levels (Figure 4I-J’’), (ii) that cells transfected with constitutively active β-catenin-GFP have higher levels of Rac1 than control untransfected cells (Figure 4K-K’’) and (iii) that Rac1 is stabilized in cells transfected with CA-Lrp6-GFP when compared to untransfected cells (Figure4L-L’’). In the original EIPA experiment in SW480 cells, now deleted from this version of the manuscript, we tested the cell viability using a Vi-Cell Beckman-Coulter Viability Analyzer and found that cells were 96-98% viable but proliferation was strongly decreased after 12 h of EIPA treatment. The effect of brief Rac1 inhibition (7 min) in decreasing dorsal development in embryos at the critical 32-cell stage is robust (Figure 4A-C). In addition, coinjection of EHT is able to entirely block the effects of microinjected xWnt8 mRNA (compare Figure 4E to 4G, see also Figure 4H), suggesting that Rac1 is required for Wnt signaling. Quantitative target gene expression analysis is provided for the embryo experiments (Figure 4C and 4H); for the stabilization of Rac1 by Wnt we are not providing quantitative measurements, but found similar results with 3 independent approaches (LiCl, CA-β-catenin and CA-Lrp6).

1. The data on FAK inhibition and TES trafficking are poorly integrated with the rest of the paper.

We attempted to better relate the TES trafficking to our previous paper showing that canonical Wnt signaling induces focal adhesion and Integrin-β1 endocytosis. We now write in the results: “We have previously reported a crosstalk between the Wnt and focal adhesion (FA) signaling pathways. Wnt3a treatment rapidly led to the endocytosis of Integrin β1 and of multiple focal adhesion proteins into MVBs (Tejeda-Muñoz et al., 2022). FAs link the actin cytoskeleton with the extracellular matrix (Figure 6A), and we now investigated whether FA activity is affected by Wnt signaling, PMA treatment and CRC progression”.

Reviewer #3 (Recommendations For The Authors):

The reliance on pleiotropic inhibitors is a weakness and should be supplemented by genetic approaches to inhibit macropinocytosis.

We agree, but that would be outside of the scope of this study.

Version 2

Abstract

Summary

Activation of the Wnt pathway lies at the core of many human cancers. Wnt and macropinocytosis are often active in the same processes, and understanding how Wnt signaling and membrane trafficking cooperate should improve our understanding of embryonic development and cancer. Here we show that a macropinocytosis activator, the tumor promoter Phorbol 12-myristate 13-acetate (PMA), enhances Wnt signaling. Experiments using the Xenopus embryo as an in vivo model showed marked cooperation between the PMA phorbol ester and Wnt signaling, which was blocked by inhibitors of macropinocytosis, Rac1 activity, and lysosome acidification. Human colorectal cancer tissue arrays and xenografts in mice showed a correlation of cancer progression with increased macropinocytosis/multivesicular body/lysosome markers and decreased GSK3 levels. The crosstalk between canonical Wnt, focal adhesions, lysosomes, and macropinocytosis suggests possible therapeutic targets for cancer progression in Wnt-driven cancers.

Introduction

Cancer progression is a process during which mutations in various signaling pathways accumulate in the cell leading to uncontrolled growth (Kinzler and Vogelstein, 1996; Hanahan and Weinberg, 2011). Cancer progression is also influenced by tumor promoters, substances that do not mutate the DNA by themselves but nevertheless facilitate cancer development. In 1941 I. Berenblum and Peyton Rous found that inflammatory agents such as croton oil or turpentine could induce skin cancer when applied repeatedly to mouse or rabbit skin, but only when an initiator mutagen such as benzopyrene or anthracene had been previously painted on that patch of skin (Berenblum, 1941; Rous and Kidd, 1941). The active agent in croton oil was later purified and found to be the phorbol ester Phorbol-12-myristate-13-acetate (PMA), also known as 12-O-tetradecanoylphorbol-13-acetate (TPA). The molecular mechanism by which PMA promotes cancer is incompletely understood but involves the activation of Protein Kinase C (PKC) in the plasma membrane, where PMA mimics the endogenous second messenger Diacylglycerol (DAG) (Nishizuka, 1984). Tumor promoters cause chronic inflammation and significantly contribute to tumorigenesis (Weinberg, 2007).

The canonical Wnt/β-catenin pathway is a major driver of cancer and plays a crucial role in embryonic development (MacDonald et al., 2009; Nusse and Clevers, 2017). In the case of colon cancer, the great majority of tumors are initiated by mutation of the tumor suppressor Adenomatous Polyposis Coli (APC), which is a component of the destruction complex that normally degrades the transcriptional regulator β-catenin, leading to its accumulation in the nucleus and benign polyp formation (Kinzler and Vogelstein, 1996; Segditsas and Tomlinson, 2006). Activation of the Wnt/β-catenin pathway is also the driving force in many other cancers (Galluzzi et al., 2019). A key regulator of Wnt signaling is Glycogen Synthetase 3 (GSK3), which phosphorylates β-catenin, leading to the formation of a phosphodegron that triggers its degradation in the proteasome. Upon activation of the Wnt receptors, GSK3, together with the destruction complex, is sequestered into multivesicular bodies (MVBs), also known as the late endosomes (Taelman et al., 2010; Vinyoles et al., 2014). The translocation of GSK3 from the cytosol into MVBs/lysosomes leads to the stabilization of many other cellular proteins in addition to β-catenin, in a process designated Wnt-STOP (Wnt-Stabilization of Proteins) (Acebron et al., 2014; Albrecht et al., 2021).

The many GSK3 phosphorylation targets stabilized by Wnt signaling include regulators of macropinocytosis such as Ras, Rac1, and Pak1 (Taelman et al., 2010; Albrecht et al., 2020; Jeong et al., 2012). A recent advance was the realization that Wnt signaling requires macropinocytosis (i.e., cell drinking) in order to traffic activated Wnt receptors and GSK3 to MVBs and lysosomes (Redelman-Sidi et al., 2018; Tejeda-Muñoz et al., 2019). Wnt-induced macropinocytosis also causes the internalization of focal adhesions and integrins from the cell surface into MVBs (Tejeda et al., 2022). The requirement of macropinocytosis for Wnt signaling is revealed, for example, by the decrease in nuclear β-catenin when colorectal cancer (CRC) cells mutant for APC were treated with macropinocytosis inhibitors such as the derivative of the diuretic Amiloride known as 5-(N-Ethyl-N-isopropyl) Amiloride (EIPA) (Tejeda-Muñoz et al., 2019; Tejeda-Muñoz and De Robertis, 2022a, 2022b;).

Macropinocytosis is emerging in the field as a promising target for cancer treatment (Lambies and Commisso, 2022). In addition, it has been found that the macropinocytosis induced by Wnt signaling causes significant lysosome acidification and activation (Albrecht et al., 2020). There is also accumulating evidence showing that the size and number of lysosomes increase during carcinogenesis (Cardone et al., 2005; Kroemer and Jäättelä, 2005; Zhitomirsky and Assaraf, 2016). Lysosomes, previously thought to be the garbage disposal of the cell, have emerged as central organelles in cancer (Kallunki et al., 2012; Kirkegaard and Jäättelä, 2009).

From the considerations above, investigating the intersection of Wnt signaling, macropinocytosis, lysosomes, focal adhesions, and membrane trafficking, is of great interest. While in most cells macropinocytosis requires stimulation of receptor tyrosine kinase (RTK) (Haigler et al., 1979; Yoshida et al., 2018), Ras (Bar-Sagi and Feramisco 1986; Commisso, 2013), or Wnt (Redelman-Sidi et al., 2018; Tejeda et al., 2019), some cells, such as amoebae and macrophages, have constitutive macropinocytosis. Importantly, Swanson discovered that the phorbol ester PMA stimulated macropinocytosis even in macrophages that were thought to have fully activated constitutive macropinocytosis (Swanson, 1989).

In the present study, we asked whether phorbol ester could interact with canonical Wnt signaling. Using the Xenopus embryo, which provides a premier model system to analyze the Wnt pathway, we found that PMA, which lacked any effects on its own, greatly sensitized the embryo to dorsalization by Wnt signaling, leading to the formation of radial head structures lacking trunks. The dorsalizing effect of PMA on embryos was blocked by the macropinocytosis inhibitor EIPA, the lysosomal V-ATPase inhibitor Bafilomycin A (Baf), or the Rac1 inhibitor EHT1864. Rac1 is an upstream regulator of the p21-activated kinase 1 (Pak1) required for actin-mediated macropinocytosis. Lysosomal activity was specifically required at the 32-cell stage for the activity of the Wnt-mimicking GSK3 inhibitor Lithium Chloride (LiCl). Ectopic axes induced by dominant-negative GSK3 (DN-GSK3-β) or β-catenin mRNAs were blocked by EIPA or dominant-negative Rab7, indicating a widespread requirement of membrane trafficking for Wnt signaling. This involvement of membrane trafficking in Wnt-driven cancer was supported by immunostaining results of arrayed human CRC histological sections, in which we found that the grade of malignancy increased Pak1, CD63 (an MVB marker), and V0a3 (a V-ATPase subunit that marks lysosomes) levels, while decreasing GSK3. In Xenopus embryos, twinning induced by Xwnt8 mRNA microinjection was blocked by inhibition of Rac1. In cultured CRC SW480 cells reconstituted with APC, the focal adhesion marker Tes was cytoplasmic, while the same cells mutant for APC (in which the Wnt pathway is activated) Tes was nuclear. The results suggest that macropinocytosis and membrane trafficking play an important role in embryogenesis and cancer progression.

Results

The Phorbol Ester PMA Potentiates Wnt-like Signals in Xenopus embryos Via Macropinocytosis and Lysosomal Acidification

Wnt is one of the earliest signals in the embryo and it induces the formation of the primary dorsal-ventral and head-to-tail axes (Loh et al., 2016; Niehrs, 2012). It is known that β-catenin is required for Wnt signaling activation on the dorsal side during early cleavage. Recently, we reported that inhibition of macropinocytosis with EIPA, or of lysosomal acidification with Baf, ventralized endogenous early dorsal axis formation, leading in particular to reduced head and brain structures (Tejeda-Munoz and De Robertis, 2022a). Activation of lysosomes on the dorsal side of the embryo was detectable as early as the 64-cell stage (Tejeda-Munoz and De Robertis, 2022a).

The tumor promoter PMA is known to increase macropinocytosis in macrophages (Swanson, 1989). To test its activity during early development, we used a sensitized system in which a small amount of the GSK3 inhibitor Lithium chloride (LiCl) (4 nl at 300 mM in a single ventral injection at 4-cell) causes only a small dorsalization with slightly enlarged heads (Kao et al., 1986; Tejeda-Munoz and De Robertis, 2022a). The dorsalizing effects of PMA were most striking when phenotypes were analyzed at the early tailbud stage (Figure 1A-H). LiCl alone expanded head structures only slightly and PMA alone was without effect (Figures 1A, B, and E). However, the combination of PMA and LiCl lead to radially dorsalized embryos consisting of head structures lacking trunks (compare Figure 1B to 1F). This striking phenotype was eliminated by co-injection of EIPA or Baf, suggesting that macropinocytosis and lysosomal acidification are required for the effect of PMA (Figure 1G and 1H). Injection of Baf alone decreased head and neural development, supporting a requirement for lysosomes for endogenous dorsal axis formation (Figure 1D). Target genes of the early Wnt signal can be detected at blastula (stage 9.5, just after the start of zygotic transcription) by measuring transcripts of Siamois and Xnr3 by qRT-PCR (Tejeda-Munoz and De Robertis, 2022a). As shown in Figure 1I, PMA alone (4 nl at 500 nM) was without any effect, LiCl increased the Wnt target genes, and the combination of PMA and LiCl synergized significantly elevated Siamois and Xnr3 transcript levels, while EIPA and Baf blocked the increase in Siamois and Xnr3 caused by PMA/LiCl (Figure 1I).

The Phorbol Ester PMA Synergizes with GSK3 Inhibition and Wnt3a and this Cooperation Requires Macropinocytosis and Lysosome Acidification (A) Uninjected control embryo at stage 24. (B) LiCl (4nl 300 mM, 1 x ventral) dorsalized the embryo only slightly. (C) EIPA (1 mM, 1x ventral) alone did not produce a distinctive phenotype at this concentration. (D) The vacuolar ATPase (V-ATPase) inhibitor Bafilomycin A1 (Baf) (incubation with 5 μM for 7 minutes at 32 cells) ventralized embryos. (E) A single injection of PMA (500 nM) did not show any phenotypic effect. (F) PMA and LiCl strongly synergized, resulting in embryos with radial heads. (G) Coinjection of the EIPA macropinocytosis inhibitor blocked dorsalization caused by LiCl plus PMA. (H) Incubation with Baf lysosomal acidification inhibitor blocked dorsalization caused by LiCl plus PMA. Number of embryos were as follows A = 70, 100%; B = 80; 98%; C = 75, 94%; D = 82, 92%; E = 76, 100%; F = 112, 98% G = 75, 93%; H = 85, 98%. Scale bar, 500 μm. (I) PMA increases the effects of LiCl in Wnt signaling in early embryos. qRT-PCR analysis at blastula stage 9.5 of the Wnt target genes Siamois and Xnr3 normalized for Ornithine decarboxylase (ODC) transcripts. Embryos received a single injection of 4 nl of LiCl (300 mM with or without PMA (500 nM). Error bars denote SEM (n ≥ 3) (**P < 0.01). (J) Luciferase assay in HEK293BR cells showing cooperation between PMA and LiCl, and its inhibition by Baf. (K-P) PMA and Wnt3a cooperated in stimulating macropinocytosis, which was sensitive to inhibition by EIPA, in 3T3 cells. TMR-dextran 70 kDa was added for 1 h. (Q) Quantification of the Wnt3a+PMA results. Experiments with cultured cells were biological triplicates. Error bars denote standard deviation (****P < 0.0001 and **P < 0.01). Also see Figure S1.

Using the BAR (β-catenin activity reporter) assay system (Biechele and Moon, 2008) in a cultured HEK293 permanent BAR/Renilla cell line, PMA synergized with LiCl when assayed by transcriptional BAR luciferase activity (Figure 1J). PMA on its own had no detectable effect on BAR reporter activity (Figure S1J). In this cell line PMA cooperated with LiCl to increase β-catenin levels (Supplementary Figure S1A-H). PMA alone caused a slight increase in β-catenin protein staining, but this lacked statistical significance (Figure S1C and S1I), presumably due to a small amount of endogenous Wnt secreted by this cell line (Colozza et al., 2020). Importantly, the effect of PMA on the activation of the Wnt pathway by GSK3 inhibition was abrogated by the inhibition of V-ATPase with Baf (Figure S1H and S1I).

To test the effect of PMA on Wnt-induced macropinocytosis we used an uptake assay in which 3T3 cells were treated with dextran for 1 h (Figure 1K-P). The macropinocytosis marker TMR-dextran 70 kDa has a hydrated diameter of more than 200 nM and is diagnostic of large endocytic cups, in particular when sensitive to the macropinocytosis inhibitor EIPA (Commisso et al., 2013, 2014). In the presence of Wnt3a, but not in the absence, PMA strongly induced dextran uptake and this effect was blocked by EIPA (Figure 1K-P, quantitatied in Figure 1Q). PMA also increased dextran uptake in the SW480 CRC cell line that has constitutive Wnt signaling, even in the absence of added Wnt (Figure S1J and K). In time-lapse video microscopy of SW480 cells with the plasma membrane tagged by myristoylated-GFP (mGFP), PMA addition triggered the abundant formation of macropinosome cups within minutes of addition when compared to DMSO controls (Supplementary Movie S1).

These results in embryos and cultured cells indicate that the tumor promoter PMA potentiates Wnt signaling and that this cooperation requires macropinocytosis and lysosomal acidification.

Lysosomal Activity is Required at the 32-cell Stage for Early Wnt Activation

In previous work, we had shown that immersion of embryos in the V-ATPase inhibitor Baf at 5 µM for only 7 minutes at the 32-cell stage resulted in ventralized microcephalic embryos lacking cement glands (Tejeda-Muñoz and De Robertis, 2022a). To test whether this ventralization is caused by inhibition of the endogenous Wnt pathway signal, we now carried out order-of-addition experiments. When an initial Baf incubation of 7 min was followed by a second brief incubation in LiCl while still at the 32-cell stage, the ventralization by Baf was rescued, and the resulting tailbud embryos had a LiCl-like dorsalized phenotype (Figure 2A-D). The initial effect of GSK3 by LiCl is to trigger macropinocytosis (Albrecht et al., 2020). The phenotypes are supported by the activation of Wnt signaling at the blastula stage, as indicated by qRT-PCR of direct Wnt target genes (Figure 2E). In the converse experiment, when embryos were first briefly treated with LiCl, washed, and subsequently incubated for 7 minutes in Baf, the effect of lysosomal inhibition was dominant, resulting in ventralized tailbud embryos (Figure 2F-2J). These order-of-addition experiments strongly support the proposal that membrane trafficking and endogenous lysosomal activity are required to generate the initial endogenous Wnt signal that takes place at the 32-cell stage of development (Tejeda-Muñoz and De Robertis, 2022a).

Order-of-addition Experiment Showing that Lysosomal Acidification at 32-cell Stage is Required for Early Wnt/β-catenin Activation (A) Untreated embryo. (B) Baf treatment (5 μM for 7 minutes) inhibited endogenous axis formation resulting in ventralized embryos lacking heads at early tailbud. (C) Embryos treated with Baf, washed, and then immersed in 300 mM LiCl an additional 7 minutes formed heads, resulting in dorsalized embryos with enlarged heads and short trunks. Note that all treatments were done at 32-cell stage, which is critical for the early Wnt signal. (D) LiCl treatment alone dorsalized embryos. (E) Quantitative RT-PCR (qPCR) for Wnt target genes Siamois and Xnr3 at blastula, confirming that the phenotypic effects are due to early activation of the Wnt pathway. (F) Untreated embryo. (G) Embryos incubated with LiCl at 32-cell were dorsalized. (H) Embryos first treated with LiCl and then subsequently with Baf resulted in ventralized embryos lacking heads. (I) Baf treatment caused ventralization and small heads. (J) Quantitative RT-PCR for Wnt target genes Siamois and Xnr3 showing that inhibiting V-ATPAse at the 32-cell stage blocks the effect of earlier LiCl treatment. The numbers of embryos analyzed were as follows: A = 115, 100% B = 124, 98% with phenotype; C = 132, 97%; D 99%; F = 132, 100%; G = 135, 99%; H = 126, 97%; I = 129, 98% (Scale bars, 500 μm.). Error bars denote SEM (n ≥ 3) (**P < 0.01).

Wnt Pathway Signaling in Microinjected Embryos or APC Mutant Cells Requires Macropinocytosis and Membrane Traficking

Next, we tested the requirement for macropinocytosis of different components of the Wnt pathway using the axis duplication assay (McMahon and Moon, 1989). Microinjection of Wnt-mimic dominant-negative GSK3 (DN-GSK3-β) mRNA, a catalytically inactive form of GSK3, induced complete axis duplications, which were blocked by the co-injection of the macropinocytosis inhibitor EIPA (1 mM, 4 nl) (Figure 3A-D). Secondary axis induction by β-catenin mRNA was also abrogated by EIPA, as well as the membrane trafficking inhibitor dominant-negative Rab7 (DN-Rab7) (Figure 3E-H). β-catenin is a transcriptional activator, and one might have not expected a requirement for membrane trafficking. However, axis formation is a multistep process and appears to require macropinocytosis and Rab7 activity in addition to transcriptional activation of Wnt target genes.

Induction of Secondary Axes by the GSK3/β-catenin Pathway Requires Membrane Trafficking (A) Control embryo at early tailbud. (B) Embryo injected with DN-GSK3 (dominant-negative GSK3-β, 150 pg 1 x ventral) mRNA, showing double axes (arrowheads). (C) Injection of EIPA 1 x ventral alone showed no phenotypic effect (4 nl,1 mM). (D) Coinjection of DN-GSK3 and EIPA blocked double axis formation. (E) Uninjected control embryo. (F) Activation of Wnt signaling via injection of β-catenin mRNA (80 pg) induced complete twinned axes (arrowheads). (G) Embryo coinjected with β-catenin mRNA and DN-Rab7 (500 pg) showing that membrane trafficking is required for secondary axis formation. (H) Coinjection of β-catenin mRNA and the macropinocytosis inhibitor EIPA blocks axial duplication. The numbers of embryos analyzed were as follows: A = 62, 100%; B = 75, 94% with double axes; C = 76, 100%; D = 70; 98%; E = 70, 100%; F = 73, 98%; G= 69, 75%; H = 82, 97%; four independent experiments; (Scale bars, 500 μm.). Error bars denote SEM (n ≥ 3) (**P < 0.01).

Macropinocytosis is an actin-driven process that is orchestrated by Pak1 kinase, which in turn is activated by Rac1. Rac1 activity is essential for the formation of lamellipodia and macropinocytic cups (Hall, 1998). EHT1864 (EHT) is a specific inhibitor of Rac1 that has the practical advantage of being water-soluble (Hampsch et al., 2017). When embryos were immersed in EHT at the critical 32-cell stage for 7 minutes (10 mM), head structures were greatly reduced, and ventral posterior structures expanded (Figures 4A and 4B). This was accompanied by a significant increase of transcripts of the ventral markers Sizzled and Vent1 at the blastula stage (Figure 4C).

Rac1 Inhibitor EHT1864 Blocks Wnt Signaling in Xenopus Embryos, and Rac1 Levels are Stabilized by Treatments that Increase Wnt/β-catenin Signaling (A) Uninjected control embryo. (B) Incubation of the Xenopus embryos with the Rac1 inhibitor EHT at 32-cell stage (7 min, 10 mM) resulted in a ventralized phenotype with a small head in the anterior (A, arrowhead) and expanded ventral structures in the posterior (P). Rac1 activity is required for macropinocytosis, see Figure S2). (C) qRT-PCR of gastrula stage embryos showing increased ventral markers Szl and Vent1 after Rac1 inhibition. (D) Control embryo. (E) Injection of Wnt mRNA (2 pg) induces complete duplicated axes (arrows). (F) Injected embryos with EHT (1 mM, 4 nl 1 x ventral) alone showed no phenotypic effect at this concentration. (G) EHT coinjected with Wnt8 mRNA blocked double axis formation. (H) qRT-PCR of Wnt target genes Siamois and Xnr3 at blastula confirming that Rac1 is required for early Wnt signaling. (I-J’’) Treatment of 3T3 cells with 40 mM LiCl increases β-catenin and Rac1 levels. (K-L’’) Transfection of 3T3 cells with stabilized constitutively active forms of β-cateninGFP or Lrp6GFP increased levels of Rac1 protein in transfected cells compared to untransfected ones. The numbers of embryos analyzed were as follows: A = 52, 100%; B = 47, 95% with ventralized small head phenotype; D = 58, 100%; E = 67; 97%; F = 64, 96%; G = 62, 97%, four independent experiments. Scale bars for embryos 500 μm; scale bars for immunofluorescence, 10 μm. qRT-PCR experiments represent biological replicates. Error bars denote SEM (n ≥ 3) (**P < 0.01). Also see Figure S2

Microinjection of a small amount of EHT (1 mM, 4 nl, 1x ventral at 4-cell) was without effect on its own but was able to block the axis-duplication effect of microinjecting Wnt8 mRNA (Figure 4D-G). Quantitative RT-PCR analyses at the blastula stage confirmed that inhibiting Rac1 strongly reduced the induction of the Wnt target genes Siamois and Xnr3 caused by xWnt8 mRNA microinjection at the late blastula stage (Figure 4H).

In SW480 CRC cells, which have constitutively activated Wnt due to mutation of APC (Leibovitz et al., 1976; Faux et al., 2004), EHT inhibited the size of cell spheroids in hanging drop cultures (Figure S2A-C) and, importantly, significantly inhibited macropinocytosis uptake of TMR-dextran in these cancer cells (Figure S2D-F).

Rac1 itself was stabilized in 3T3 cells by Wnt signaling, as indicated by experiments showing (i) that LiCl treatment increased levels of Rac1 protein (Figure 4I-J’’), (ii) that cells transfected with constitutively active β-catenin-GFP had higher levels of Rac1 than control untransfected cells (Figure 4K-K’’) and (iii) that Rac1 immunostaining was stabilized in cells transfected with CA-Lrp6-GFP when compared to untransfected cells (Figure4L-L’’).

The results show that signaling by the Wnt/β-catenin pathway requires macropinocytosis, Rac1, and membrane trafficking in Xenopus embryos and cancer cells.

Tumor Progression in Human Colorectal Cancers Correlates with a Role for membrane Traficking in Wnt/β-catenin Signaling

Colorectal cancer is particularly favorable for the analysis of tumor progression because it gradually accumulates mutations, and in over 85% of the cases the initial driver mutation is in the APC tumor suppressor (Kinzler and Volgelstein, 1996; Weinberg, 2007). This offered an opportunity to test our proposal that membrane trafficking and its molecular components play an important role in Wnt-driven cancers (Albrecht et al., 2021). Commercially available arrays of paraffin-sectioned histological sections containing 90 cases of adenocarcinoma of various grades I-IV and 90 samples of corresponding adjacent normal tissues (from TissueArray) were used for immunostaining. Arrays were double stained for β-catenin and antibodies against proteins involved in the macropinocytosis/MVB/lysosome/GSK3 pathway. Each individual section was evaluated, and normal colon, adenocarcinoma I, and adenocarcinoma IV were compared to assess the effects of cancer progression.

As shown in Figure 5, β-catenin levels consistently correlated with grade IV malignancy when compared to normal colon or grade I cancers. Pak1, a kinase required for macropinocytosis and Wnt signaling (Redelman-Sidi et al., 2018; Albrecht et al.,2020) was significantly elevated with increased malignancy (Figure 5A-C’’). CD63, a tetraspan protein that marks MVB intraluminal vesicles (Escola et al., 1998), reached high levels in adenocarcinoma IV (Figure 5D-F’’), particularly in cells with the highest β-catenin expression (compare Figure 5F to 5F’). This increase in MVBs is in agreement with the GSK3 sequestration model of Wnt signaling (Taelman et al., 2010). The V-ATPase subunit V0a3, a marker for acidic lysosomes (Ramirez et al., 2019), also increased strongly with the CRC malignancy (compare Figure 5G to 5I). Importantly, the cellular levels of total GSK3 decreased with malignancy (Figure 5J-L’’, compare 5J to 5L), as predicted by the sequestration model (Taelman et al., 2010). During Wnt signaling, GSK3 is sequestered into MVBs which are trafficked into lysosomes and a decrease was to be expected after sustained Wnt activation. Previously, we had been unable to demonstrate a decrease in total cellular GSK3 levels during acute Wnt signaling experiments (Taelman et al., 2010). The co-localization of β-catenin with Pak1, CD63, V0a3, or GSK3 in human colorectal adenocarcinomas was quantified in Figures S3A-S3D.

Degree of Human CRC Malignancy is Associated with Increased Macropinocytosis/MVB/lysosome markers and decreased GSK3 levels (A-A’) Normal human colon paraffin section stained with the macropinocytosis marker Pak1 and β-catenin, respectively. (A’’) Merged image with DAPI, a few cells colocalize both markers. (B-B’’) Pak1 and β-catenin levels are moderately increased at an early stage I CRC adenocarcinoma. (C’-C’’) Strong colocalization between Pak1 and β-catenin was observed in advanced stage IV CRC (inset). (D-D’’) Normal human colon section stained with the MVB marker CD63 and β-catenin; colocalization was not observed. (E-E’’) CD63 and β-catenin were stabilized in adenocarcinoma I, and moderate colocalization was found between CD63 and β-catenin. (F-F’) CD63 and β-catenin were strongly stabilized and colocalized in adenocarcinoma stage IV CRC. (F’’) Merge; note the striking colocalization between the MVB marker CD63 and β-Catenin in advanced stages of cancer (inset). (G-G’’) Normal colon stained for V0a3 (a subunit of V-ATPase that marks lysosomes) and β-catenin. (H-H’’) Stage I adenocarcinoma with moderately increased levels of lysosomes and β-catenin. (I-I’’) Strong co-localization of lysosomes and β-catenin in stage IV CRC (see inset). (J-J’’) Human colon array stained with the GSK3 and β-catenin. (K-K’’) GSK3 decreases, and β-catenin increases, at early stages of carcinogenesis, (L-L’’) GSK3 levels are very low in advanced CRC compared to normal human colon, while β-catenin levels are very high. Scale bars, 10 μm. Also see Figure S3 for quantifications and mouse xenografts of CRC cells.

These results were confirmed by examining immunostaining levels of Pak1, V0a3, and GSK3 compared to β-catenin protein in mouse xenografts in a CD1 NU/NU nude mouse model injected with SW480 human CRC cells (in which Wnt activation is known to be caused by APC mutation, Faux et al., 2004). In this case, tumor cells were compared to normal colon mouse tissue sections. The results showed that CRC cells had high levels of β-catenin, Pak1, and lysosomes (Figure S3E-L), while GSK3 levels were reduced (compare Figure S3K to S3L) in agreement with the results in human tissue arrays.

These in vivo observations in human tumors are consistent with the view that macropinocytosis components, MVBs, and lysosomes increase with CRC malignancy. Further, the decrease in total GSK3 levels suggests a role in the trafficking and degradation of this key enzyme in cancer.

Wnt Affects the Subcellular Localization of Focal Adhesion Components

We have previously reported a crosstalk between the Wnt and focal adhesion (FA) signaling pathways in which Wnt3a treatment rapidly led to the endocytosis of Integrin β1 and of multiple focal adhesion proteins into MVBs (Tejeda-Muñoz et al., 2022). FAs link the actin cytoskeleton with the extracellular matrix (Figure 6A), and we now investigated whether FA activity is affected by Wnt signaling, PMA treatment or CRC progression. FA components containing LIM domains can shuttle between the nucleus and cytoplasm (Nix and Beckerle, 1997; Kadrmas and Beckerle, 2004; Anderson et al., 2021). We examined whether constitutive Wnt signaling regulates FA trafficking using the SW480 and SW480APC system, in which the stable reconstitution with full-length APC restores these CRC cells to a non-malignant phenotype (Faux et al., 2004), which is accompanied by a decrease in β-catenin levels (Figure 6B). As shown in Figure 6C to 6D’’, the PET-LIM domain FA protein Tes (also known as Testin) (Garvalov et al., 2003) was cytoplasmic in SW480APC, but was nuclear in SW480 cells in which Wnt is activated by APC mutation. The nuclear localization of Tes in SW480 cells required macropinocytosis, as suggested by treatment with EIPA which resulted in its re-localization to the cytoplasm (Figure 6E-6F’’). In the nude mouse xenograft model, Tes was abundant in the cytoplasm of normal colon epielium, while in transplanted SW480 cells it colocalized with β-catenin in the nucleus of those cells with the highest β-catenin levels (Figure 6G-6I’’). The function of TES is poorly understood, but it has been proposed to act as a tumor suppressor (Tatarelli et al., 2000; Tobias et al., 2001).

A Focal Adhesion Protein Changes Its Nucleocytoplasmic Distribution After Wnt Signaling Activation, and Focal Adhesion Kinase Inhibition Affects Wnt/β Signaling (A) Diagram showing how focal adhesions connect the actin cytoskeleton to the extracellular matrix via integrins; Wnt signaling changes the distribution of the focal adhesion protein Tes from focal adhesion sites to the nucleus. (B) Western blot of SW480 cells with and without stable expression of APC; note the reduction in Tes expression and increase in β-catenin levels in SW480 cells lacking full-length APC. GADPH was used as loading control. (C-C’) Colon cancer cell line SW480 stably restored with full-length APC stained with Tes and β-catenin antibodies. Note that in SW480APC cells TES is absent from the nucleus (inset). (D-D’’) In SW480 CRC cells both Tes and β-catenin are located the nucleus (inset). (E-E’’) Colon cancer cell line SW480 stained with Tes (in green) and β-catenin (in red) antibodies as a control. (F-F’’) Inhibiting macropinocytosis with EIPA (40 μM) treatment restored the cytoplasmic localization of the focal adhesion protein Tes in SW480 cells and strongly inhibited β-catenin levels. (G) Quantification of colocalization between Tes and β-catenin in normal mouse colon and SW480 xenografts. (H-H’) Immunohistochemistry of the CD1 NU/NU normal mouse colon showing Tes and β-catenin distribution. (H’’) Merge showing modest co-localization (inset). (I-I’’) Immunohistochemistry images of the CD1 NU/NU mouse xenograft model showing nuclear TES, which is strongly colocalized with β-catenin levels in transplanted human SW480 cancer cells (inset). Scale bars, 10 μm. (J) Reporter assay in HEK293T cells stably expressing BAR and Renilla, showing strong inhibition of the β-catenin transcriptional activity induced by LiCl (40 mM) by FAK inhibitor PF-00562271 (20 μM) after overnight treatment. Error bars denote SEM (n ≥ 3) (**P < 0.01). (K-L’’’) Immersion of Xenopus embryos in FAK inhibitor (7 min at 32-cell stage) inhibits head and dorsal development. Bright field and several angles of in situ hybridization with pan-neural marker Sox2 are shown to demonstrate reduction in brain structures indicated by the arrowhead; scale bar 500 μm.

Focal Adhesion Kinase (FAK) is a key regulator of FA maintenance and signaling, known to interact with the Wnt pathway in various ways (Chuang et al., 2022). In HEK293 BAR/Renilla cells, treatment with the FAK inhibitor PF-00562271 strongly decreased β-catenin signaling by LiCl inhibition of GSK3 (Figure 6J). In the context on the Xenopus embryo, treatment with the FAK inhibitor (100 µM, 7 min at 32-cell stage) resulted in ventralized embryos with reduced dorsal-anterior structures such as central nervous system, in particular brain structures stained by the pan-neural marker Sox2 (Figure 6K’-L’’). The addition of the tumor promoter PMA to SW480 cells (in which the Wnt pathway is constitutively activated) strongly increased protein levels of β-catenin and FAK protein (Figure S4C-D’’). The effect of PMA on β-catenin and FAK levels was abrogated by EIPA or Baf treatment (Figure S4C-S4H’’), suggesting that it requires macropinocytosis and lysosome activity.

Taken together, the results suggest a possible involvement of focal adhesion components and FAK in Wnt signaling, in particular during the endogenous early Wnt signal in Xenopus embryos. The results also suggest that nuclear Tes could serve as a Wnt-regulated marker in CRC. The regulated shuttling of FA components between nucleus and cytoplasm is probably just the tip of the iceberg of the role of membrane trafficking in Wnt-driven cancers.

Discussion

The present work was inspired by the discovery that PMA increases macropinocytosis (Swanson, 1989). PMA plays an important role in our understanding of tumor biology because it is the archetypal tumor promoter, able to promote tumor growth without further mutation of the DNA (Berenblum, 1941; Weinberg, 2007). Our experiments support the role of membrane trafficking in canonical Wnt signaling in embryos and cancer cells and are summarized in the model in Figure 7. The main finding was a synergistic positive effect between the tumor promoter phorbol ester (PMA) and activation of the Wnt/β-catenin pathway that was blocked by EIPA or Baf (Figure 1). EIPA is a macropinocytosis inhibitor of the Na+/H+ exchanger at the plasma membrane; it causes acidification of the submembranous cytoplasm and prevents actin polymerization (Koivusalo et al., 2010). Bafilomycin A (Baf) is an inhibitor of the V-ATPase that drives membrane trafficking and lysosomal acidification. PMA activates the PKC pathway (Takai et al., 1979; Nishizuka, 1984), and there is previous literature on the crosstalk between the PKC and Wnt pathways (Schwarz et al., 2013; Goode et al., 1992). What is novel in the present paper is the crosstalk between macropinocytosis, membrane trafficking, lysosomes, focal adhesions, PMA, and Wnt signaling.

Model of PMA Synergy with Wnt Signaling through Macropinocytosis and Membrane Trafficking Wnt and PMA are activators of macropinocytosis, which is a cell drinking process driven by an actin meshwork that requires the small GTPase Rac1. Membrane trafficking into MVBs/lysosomes (marked by CD63) requires acidification by V-ATPase, which is inhibited by Bafilomycin A (Baf). Phorbol esters such as PMA stimulate macropinocytosis and synergize with Wnt signaling. Inhibiting macropinocytosis with the membrane trafficking inhibitors EIPA, Baf, or the Rac1 inhibitor EHT1864 block Wnt signaling and its cooperation with the tumor promoter PMA.

Macropinocytosis, Rac1, and Wnt signaling

Wnt signaling triggers macropinocytosis and lysosomal acidification (Albrecht et al., 2020). In cultured cells, Wnt3a-induced macropinocytosis was increased by PMA and blocked by EIPA. In embryo microinjections, Baf blocked the synergy between PMA and the Wnt-mimic LiCl. Baf also blocked development of the endogenous axis, particularly the head region when embryos were briefly immersed in Baf at the 32-cell stage, as previously reported (Tejeda-Muñoz and De Robertis, 2022a). Using order-of-addition experiments in which embryos were treated with 7 min pulses of Baf or the GSK3 inhibitor LiCl, we showed that the requirement for lysosome acidification was due to inhibition of the Wnt pathway at the crucial 32-cell stage (Figure 2). Co-injection experiments of DN-GSK3 or β-catenin mRNAs with DN-Rab7 or EIPA supported the view that membrane trafficking and macropinocytosis are essential for the induction of secondary axes in Xenopus (Figure 3).

Rac1 is an upstream regulator of the Pak1 kinase required for the actin machinery that drives macropinocytosis (Redelman-Sidi et al., 2018). Rac1 is a member of the family of small Rho GTPases that drives the formation of lamellipodia, membrane ruffles, and macropinocytosis (Hall, 1998; Egami et al., 2014; Fujii et al., 2013). The Rac1 inhibitor EHT1864 (Hampsch et al., 2017) strongly ventralized the endogenous axis of embryos after brief incubation at the critical 32-cell stage (Figure 4). In co-injection experiments EHT blocked the axis-inducing effect of microinjected xWnt8 mRNA. Rac1 is a regulator of multiple signaling cascades and has been reported to increase the nuclear translocation of β-catenin (Wu et al., 2008). Other work has shown that Rac1 binds to β-catenin and promotes the formation of nuclear β-catenin/LEF1 complexes (Jamieson et al., 2015). In the context of the present study, Rac1 protein levels were increased by treatments that enhanced Wnt/β-catenin signaling. The results suggest that Rac1 may crosstalk with Wnt through macropinocytosis in early development.

Lysosomes, cancer and focal adhesions

The lysosome is considered a promising target for the treatment of cancer (Fehrenbacher and Jäättelä, 2005; Lawrence and Zoncu, 2019). We used human CRC tissue arrays to analyze whether lysosomes and membrane trafficking correlate with cancer progression. Most colorectal tumors are initiated by mutation of APC, which drives Wnt signaling (Kinzler and Vogelstein, 1996; Segditsas and Tomlinson, 2006). It was found that advanced grades of colon carcinoma correlated with increased immunostaining for Pak1 (a driver of macropinocytosis), CD63 (a MVB/late endosome marker), and V0a3 (a lysosomal V-ATPase marker) (Figure 5). Importantly, cancer malignancy correlated with low levels of GSK3 staining, both in human samples and nude mouse SW480 xenografts. Cancer cells require lysosome function and have changes in lysosomal volume and subcellular localization during oncogenic transformation (Kallunki et al., 2012; Kirkegaard and Jäättelä, 2009; Gocheva and Joyce, 2007; Kolwijck et al., 2010). V-ATPase has been associated with acidification of the extracellular milieu in tumors (Capecci and Forgac, 2013; Hinton et al., 2009; Perona and Serrano, 1988). Extracellular tumor acidification is probably due to increased numbers of lysosomes which are exocytosed, since V0a3 was located within the cytoplasm in advanced cancer or xenografts in mice (Figures 5I and S3I). The results reported here support the view (Taelman et al., 2010; Tejeda-Muñoz and De Robertis, 2022a) that Wnt-driven lysosomal acidification and GSK3 sequestration may play a role in cancer. In addition, MVB/lysosome components could serve as markers for advanced Wnt-driven cancers.

The cytoskeleton interacts with proteins of the extracellular matrix though focal adhesions (Geiger et al., 2001). We previously reported that Wnt causes the endocytosis of FA proteins and depletion of Integrin β1 from the cell surface (Tejeda-Muñoz et al., 2022). We now report that Tes, a member of the PET-LIM family of focal adhesion proteins, changed subcellular localization after Wnt activation in SW480 cells. Other focal adhesion proteins shuttle between the nucleus, cytoplasm and focal adhesion sites, and may play important roles in cancer (Chaturvedi et al., 2012; Nix and Beckerle, 1997). FAK is a key signaling kinase that regulates focal adhesion signaling. In SW480 cells (which have constitutive Wnt signaling) FAK protein levels was increased by PMA, and this was inhibited by EIPA or Baf. The FAK inhibitor PF-00562271 inhibited β-catenin signaling in Luciferase assays, and reduced dorsal-anterior structures such as CNS in embryos after incubation for 7 min at 32-cell stage that is critical for the early Wnt signaling. These results suggest a possible crosstalk between focal adhesions, Wnt signaling and tumor promoters in cancer progression and embryonic development.

Cooperation between PMA and Wnt

PMA activates PKC signaling by mimicking its physiological second messenger diacylglycerol (DAG), an activator of protein kinase C. There is previous literature implicating PKC isozymes in the regulation of the Wnt signaling pathway (Fang et al., 2002; Goode et al., 1992) and that different PKC isoforms can have positive or negative effects on Wnt/β-catenin signaling (Ohno and Nishizuka, 2002). However, the molecular role of PKC isoforms in the Wnt signaling pathway remains poorly understood (Bhatia and Spiegelman, 2005; Schwarz et al., 2013). It has been reported that PMA can stabilize CK1ε (Casein Kinase 1 epsilon), enhance its kinase activity, and induce LRP6 phosphorylation at Thr1479 and Ser1490, leading to the activation of the Wnt/β-catenin pathway (Su et al., 2018). In the context of the present study, it is relevant that Wnt signaling rapidly triggers macropinocytosis and trafficking into lysosomes (Tejeda-Muñoz et al., 2019). Macropinocytosis requires the activation of Phosphoinositide 3 Kinase (PI3K) which leads to the formation of patches of PIP3 in the inner plasma membrane leaflet that initiate the formation of macropinocytic cups (Egami et al., 2014; Yoshida et al., 2018). During macropinosome formation, the PIP3 lipid is then converted into DAG by Phospholipase C gamma 1 (PLCγ1), and DAG activates PKC signaling (Yoshida et al., 2018). It seems likely that PMA may facilitate Wnt/β-catenin signaling by mimicking the DAG lipid in macropinosomes. We have tested this hypothesis by direct addition of DAG to cells and found cooperation with Wnt signaling (Azbazdar et. al., manuscript submitted).

In the future, it will be interesting to study the role of chronic inflammation on macropinocytosis and tumor promotion. Active PKC phosphorylates many targets, including IκB Kinase (IKK), which becomes activated and phosphorylates Inhibitor of NF-κB (IκB), targeting it for degradation and triggering inflammation (Weinberg, 2007). Tumor progression is promoted by many chronic inflammatory agents such as turpentine, distilled alcoholic beverages, hepatitis viruses, and Helicobacter pylori that do not cause mutations by themselves but regulate the expansion of previously mutant clones (Rous and Kidd, 1941; Weinberg, 2007). Taken together, the results presented here suggest that macropinocytosis, lysosomal activity, and membrane trafficking are possible therapeutic targets for tumor progression in Wnt-driven cancers.